# Binomial Distributions

{{< include _r_setup.qmd >}}



::: {#exm-binomial-stock}
Consider an extremely simplified model for the daily closing price of a certain stock.
Every day the price either goes up or goes down, and the movements are independent from day-to-day.
Assume that the probability that the stock price goes up on any single day is 0.25.
Let $X$ be the number of days in which the price goes up in the next 5 days.
:::

1.  Compute and interpret $\text{P}(X=0)$.\
    \
    \
    \
    \
    \
2.  Compute the probability that the price goes up on the first day and then down on the following four days.\
    \
    \
    \
    \
    \
3.  Why is $\text{P}(X=1)$ different from the probability in the previous part? Compute and interpret $\text{P}(X=1)$.\
    \
    \
    \
    \
4.  Suggest a general formula for the probability mass function of $X$.\
    \
    \
    \
    \

-   A discrete random variable $X$ has a **Binomial distribution** with parameters $n$, a nonnegative integer, and $p\in[0, 1]$ if its probability mass function is \begin{align*}
    p_{X}(x) & = \binom{n}{x} p^x (1-p)^{n-x}, & x=0, 1, 2, \ldots, n
    \end{align*}
-   The *binomial coefficient* (read "$n$ choose $x$") $$
      \binom{n}{x} = \frac{n!}{x!(n-x)!}
      $$ counts the number of success/failure sequences of length $n$ in which there are exactly $x$ successes. (Remember: by definition $0! = 1$.)

::: {#exm-binomial-stock-again}
Continuing @exm-binomial-stock.
:::

1.  Construct a table, plot, and spinner representing the distribution of $X$.\
    \
    \
    \
    \
2.  Compute and interpret $\text{P}(X \le 2)$.\
    \
    \
    \
    \

| $x$ |                             $p(x)$ |    Value |
|-----|-----------------------------------:|---------:|
| 0   | $\binom{5}{0}0.25^0(1-0.25)^{5-0}$ | 0.237305 |
| 1   | $\binom{5}{1}0.25^1(1-0.25)^{5-1}$ | 0.395508 |
| 2   | $\binom{5}{2}0.25^2(1-0.25)^{5-2}$ | 0.263672 |
| 3   | $\binom{5}{3}0.25^3(1-0.25)^{5-3}$ | 0.087891 |
| 4   | $\binom{5}{4}0.25^4(1-0.25)^{5-4}$ | 0.014648 |
| 5   | $\binom{5}{5}0.25^5(1-0.25)^{5-5}$ | 0.000977 |

: Table representing the Binomial(5, 0.25) probability mass function.

```{r}
#| include: false
#| eval: false

dbinom(1, 5, 0.25)
```

```{r}
#| include: false
#| eval: false

x <- 0:5

p_x <- dbinom(x, 5, 0.25)

data.frame(x, p_x) |>
  kbl(digits = 6) |>
  kable_styling(fixed_thead = TRUE)
```

```{r}
#| include: false
#| eval: false

sum(dbinom(0:2, 5, 0.25))
```

```{r}
#| include: false
#| eval: false

pbinom(2, 5, 0.25)
```

```{r}
#| include: false
#| eval: false

rbinom(10, 5, 0.25)
```

```{r}
#| label: fig-binomial-stock-sim
#| fig-cap: "Simulated distribution of $X$ with a Binomial(5, 0.25) distribution."
#| include: false
#| eval: false

N_rep <- 10000

x <- rbinom(N_rep, 5, 0.25)

# Summarize the simulated values
plot(table(x) / N_rep, type = "h", xlab = "x", ylab = "Approximate P(X = x)")
```

```{r}
#| label: spinner-binomial-stock
#| echo: false
#| fig-cap: "Spinner corresponding to the Binomial(5, 0.25) distribution (4 and 5 lumped into 4+)."

x <- c(0:3, "4+")
p <- c(dbinom(0:3, 5, 0.25), 1 - pbinom(3, 5, 0.25))

xp <- data.frame(x, p)

cdf <- c(0, cumsum(xp$p))

plotp <- (cdf[-1] + cdf[-length(cdf)]) / 2

ggplot(xp, aes(x = "", y = p, fill = x)) +
  geom_bar(width = 1, stat = "identity", color = "black", fill = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = plotp, labels = xp$x) +
  theme(axis.text.x = element_text(size = 14, face = "bold")) +
  # plot the probabilities as percents inside
  geom_text(
    aes(y = plotp, label = percent(p)),
    size = 4,
    color = c(rep("black", 4), rep("white", 1))
  ) +
  ggtitle(paste("Binomial(5, 0.25) spinner", sep = ""))
```

```{r}
#| include: false
#| eval: false

sum(dbinom(0:2, 5, 0.25))

pbinom(2, 5, 0.25)
```


::: {#exm-binomial-stock2}
Continuing @exm-binomial-stock.
:::

1.  What does the random variable $5-X$ represent? What is its distribution?\
    \
    \
    \
    \
2.  Suppose that the price is currently \$100 and each it either moves up \$2 or down \$2. Let $S$ be the stock price after 5 days. How does $S$ relate to $X$? Does $S$ have a Binomial distribution?\
    \
    \
    \
    \
3.  Recall that $X$ is the number of days on which the price goes up in the next five days. Suppose that $Y$ is the number of days on which the price goes up in the ten days after that (days 6-15). What is the distribution of $X+Y$? (Continue to assume independence between days, with probability 0.25 of an up movement on any day.)\
    \
    \
    \
    \

-   Imagine a box containing tickets
    -   Each ticket is labeled either 1 ("success") or 0 ("failure")
    -   $p$ is the proportion of tickets in the box labeled 1 ("success"); the rest are labeled 0 ("failure").
    -   Randomly select $n$ tickets from the box *with replacement* and let $X$ be the number of tickets in the sample that are labeled 1.
    -   Then $X$ has a Binomial($n$, $p$) distribution.
    -   Since the tickets are labeled 1 and 0, the random variable $X$ which counts the number of successes is equal to the sum of the 1/0 values on the tickets.
-   The above situation involves a sequence of **Bernoulli(**$p$) trials.
    -   There are only two possible outcomes, "success" (1) and "failure" (0), on each trial.
    -   The unconditional/marginal probability of success is the same on every trial, and equal to $p$
    -   The trials are independent
        -   If sampling with replacement, or
        -   If sampling without replacement but the population size (number of tickets in the box) is much larger than the sample size $n$
-   If $X$ counts the number of successes in a *fixed number*, $n$, of Bernoulli($p$) trials then $X$ has a Binomial($n, p$) distribution.

::: {#exm-binomial-situation}
In each of the following situations determine whether or not $X$ has a Binomial distribution.
If so, specify $n$ and $p$.
If not, explain why not.
:::

1.  Roll a die 20 times; $X$ is the number of times the die lands on an even number.\
    \
    \
    \
    \
2.  Roll a die 20 times; $X$ is the number of times the die lands on 6.\
    \
    \
    \
    \
3.  Roll a die until it lands on 6; $X$ is the total number of rolls.\
    \
    \
    \
    \
4.  Roll a die 20 times; $X$ is the sum of the numbers rolled.\
    \
    \
    \
    \
5.  Shuffle a standard deck of 52 cards (13 hearts, 39 other cards) and deal 5 *without replacement*; $X$ is the number of hearts dealt. (Hint: be careful about why.)\
    \
    \
    \
    \
6.  Roll a fair six-sided die 10 times and a fair four-sided die 10 times; $X$ is the number of 3s rolled (out of 20).\
    \
    \
    \
    \
7.  Randomly select a sample of 35 Cal Poly students; $X$ is the number of students in the sample who are CA residents.\
    \
    \
    \
    \



::: {#exm-binomial-stock-ev}
Continuing @exm-binomial-stock.
Consider an extremely simplified model for the daily closing price of a certain stock.
Every day the price either goes up or goes down, and the movements are independent from day-to-day.
Assume that the probability that the stock price goes up on any single day is 0.25.
Let $X$ be the number of days in which the price goes up in the next 5 days.
:::

1.  Suggest a shortcut formula for $\text{E}(X)$.\
    \
    \
    \
    \
2.  Compute $\text{E}(X)$ using the distribution of $X$ from @exm-binomial-stock. Did the shortcut formula work?\
    \
    \
    \
    \
3.  Interpret $\text{E}(X)$.\
    \
    \
    \
    \

-   If $X$ has a Binomial($n$, $p$) distribution then $\text{E}(X) = np$.


::: {#exm-dd-same-distribution}
Donny Dont is thoroughly confused about the distinction between a random variable and its distribution.
Help him understand by by providing a simple concrete example of *two different random variables* $X$ and $Y$ that have the *same distribution*.
Can you think of $X$ and $Y$ that have the same distribution but for which $\text{P}(X = Y) = 0$?
(Hint: think coin flipping.)
:::

\
\
\
\
\
