# Conditional Distributions

{{< include _r_setup.qmd >}}


-   The *joint distribution* of random variables $X$ and $Y$ is a probability distribution on $(x, y)$ *pairs*, and describes how the values of $X$ and $Y$ vary together or jointly.
-   We can also study *conditional distributions* of random variables given the values of some random variables.
    How does the distribution of $Y$ change for different values of $X$ (and vice versa)?

::: {#exm-dice-conditional}
Roll a fair four-sided die twice.
Let $X$ be the sum of the two rolls, and let $Y$ be the larger of the two rolls (or the common value if a tie).
We have previously found the joint and marginal distributions of $X$ and $Y$, displayed in the two-way table below.
:::

| $p_{X, Y}(x, y)$ |      |      |      |      |            |
|------------------|-----:|-----:|-----:|-----:|-----------:|
| $x$ \\ $y$       |    1 |    2 |    3 |    4 | $p_{X}(x)$ |
| 2                | 1/16 |    0 |    0 |    0 |       1/16 |
| 3                |    0 | 2/16 |    0 |    0 |       2/16 |
| 4                |    0 | 1/16 | 2/16 |    0 |       3/16 |
| 5                |    0 |    0 | 2/16 | 2/16 |       4/16 |
| 6                |    0 |    0 | 1/16 | 2/16 |       3/16 |
| 7                |    0 |    0 |    0 | 2/16 |       2/16 |
| 8                |    0 |    0 |    0 | 1/16 |       1/16 |
| $p_Y(y)$         | 1/16 | 3/16 | 5/16 | 7/16 |            |

1.  Compute $p_{X|Y}(6|4) = \text{P}(X=6|Y=4)$.\
    \
    \
    \
    \
2.  Construct a table, plot, and spinner to represent the conditional distribution of $X$ given $Y=4$.\
    \
    \
    \
    \
3.  Construct a table, plot, and spinner to represent the conditional distribution of $X$ given $Y=3$.\
    \
    \
    \
    \
4.  Construct a table, plot, and spinner to represent the conditional distribution of $X$ given $Y=2$.\
    \
    \
    \
    \
5.  Construct a table, plot, and spinner to represent the conditional distribution of $X$ given $Y=1$.\
    \
    \
    \
    \
6.  Compute $p_{Y|X}(4|6) = \text{P}(Y=4|X=6)$.\
    \
    \
    \
    \
7.  Construct a table, plot, and spinner to represent the distribution of $Y$ given $X=6$.\
    \
    \
    \
    \
8.  Construct a table, plot, and spinner to represent the distribution of $Y$ given $X=5$.\
    \
    \
    \
    \
9.  Construct a table, plot, and spinner to represent the distribution of $Y$ given $X=4$.\
    \
    \
    \
    \

-   The **conditional distribution of** $Y$ given $X=x$ is the distribution of $Y$ values over only those outcomes for which $X=x$. It is a distribution on values of $Y$ only; treat $x$ as a fixed constant when conditioning on the event $\{X=x\}$.
-   Conditional distributions can be obtained from a joint distribution by *slicing and renormalizing*. The conditional distribution of $Y$ given $X=x$, where $x$ represents a particular number, can be thought of as:
    -   the slice of the joint distribution corresponding to $X=x$, a distribution on values of $Y$ alone with $X=x$ fixed
    -   renormalized so that the slice accounts for 100% of the probability over the values of $Y$
-   The *shape* of the conditional distribution of $Y$ given $X=$ is determined by the shape of the slice of the joint distribution over values of $Y$ for the fixed $x$.
-   For each fixed $x$, the conditional distribution of $Y$ given $X=x$ is a different distribution on values of the random variable $Y$. There is not one "conditional distribution of $Y$ given $X$", but rather a *family of conditional distributions of* $Y$ given different values of $X$.
-   Each conditional distribution is a distribution, so we can summarize its characteristics like mean and standard deviation. The conditional mean and standard deviation of $Y$ given $X=x$ represent, respectively, the long run average and variability of values of $Y$ over only $(X, Y)$ pairs with $X=x$.
-   Since each value of $x$ typically corresponds to a different conditional distribution of $Y$ given $X=x$, the conditional mean and standard deviation will typically be functions of $x$.






```{r}
#| label: fig-dice-conditional-plots
#| echo: false
#| fig-cap: "Impulse plots representing the family of conditional distributions of $X$ given $Y$ for the dice rolling example. Each plot represents a conditional distribution of $X$ given $Y=y$ for a particular value of $y= 1, 2, 3, 4$."

x = c(2, 3, 4, 4, 5, 6, 5, 6, 7, 8)
y = c(1, 2, 2, 3, 3, 3, 4, 4, 4, 4)
p = c(1, 2 / 3, 1 / 3, 2 / 5, 2 / 5, 1 / 5, 2 / 7, 2 / 7, 2 / 7, 1 / 7)

facet_labels <- setNames(
  paste("Conditional distribution of X given Y = ", sort(unique(y)), sep = ""),
  sort(unique(y))
)

ggplot(data.frame(y, p),
       aes(x = x,
           xend = x,
           y = 0,
           yend = p)) +
  geom_segment(linewidth = 1.2, col = "skyblue") +
  # scale_color_manual(values = c("#56B4E9", "#E69F00", "#999999")) +
  scale_x_continuous(breaks = 2:8, limits = c(2, 8)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "x",
       y = "P(X = x | Y = y)") +
  facet_wrap(~y, scales= "free", labeller = as_labeller(facet_labels))

```



```{r}
#| label: fig-dice-conditional-spinners
#| echo: false
#| fig-cap: "Spinners representing the family of conditional distributions of $X$ given $Y$ in the dice rolling example. Each spinner represents a conditional distribution of $X$ given $Y=y$ for a particular value of $y= 1, 2, 3, 4$."

make_discrete_spinner <- function(x, p, title = ""){
  xp <- data.frame(x, p)
  cdf = c(0, cumsum(xp$p))
  plotp = (cdf[-1] + cdf[-length(cdf)]) / 2
  ggplot(xp, aes(x="", y=p, fill=x))+
    geom_bar(width = 1, stat = "identity", color="black", fill="white") + 
    coord_polar("y", start=0) +
    theme_void() +
    # plot the possible values on the outside
    scale_y_continuous(breaks = plotp, labels=xp$x, limits = c(0, 1)) +
    theme(axis.text.x=element_text(size=16, face="bold")) +
    # plot the probabilities as percents inside
    geom_text(aes(y = plotp,
                  label = format(percent(round(p, 3)),0)), size=4) +
    labs(title = title)
}



spinner1 = make_discrete_spinner(2, 1, "Distribution of X given Y = 1")

spinner2 = make_discrete_spinner(3:4, c(2, 1) / 3, "Distribution of X given Y = 2")

spinner3 = make_discrete_spinner(4:6, c(2, 2, 1) / 5, "Distribution of X given Y = 3")

spinner4 = make_discrete_spinner(5:8, c(2, 2, 2, 1) / 7, "Distribution of X given Y = 4")

grid.arrange(spinner1, spinner2, spinner3, spinner4, ncol = 2)

```



::: {#exm-dice-conditional-sim}
We have already discussed two ways for simulating an $(X, Y)$ pair in the dice rolling example: simulate a pair of rolls and measure $X$ (sum) and $Y$ (max), or spin the joint distribution spinner for $(X, Y)$ once.
:::

1.  Now describe another way for simulating an $(X, Y)$ pair using the spinners in @exm-dice-conditional. (Hint: you'll need one more spinner in addition to the four from the previous example.)\
    \
    \
    \
    \
2.  Describe in detail how you can simulate $(X, Y)$ pairs and use the results to approximate $\text{P}(X = 6 | Y = 4)$.\
    \
    \
    \
    \
3.  Describe in detail how you can simulate $(X, Y)$ pairs and use the results to approximate the conditional distribution of $X$ given $Y = 4$.\
    \
    \
    \
    \
4.  Describe in detail how you can simulate values from the conditional distribution of $X$ given $Y=4$ without simulating $(X, Y)$ pairs.\
    \
    \
    \
    \


-   Rather than directly simulating from a joint distribution, we can simulate an $(X, Y)$ pair in two stages:
    -   Simulate a value of $X$ from its marginal distribution. Call the simulated value $x$.
    -   Given $x$, simulate a value of $Y$ from the conditional distribution of $Y$ given $X = x$. There will be a different distribution (spinner) for each possible value of $x$.
-   This "marginal then conditional" process is essentially implementing the multiplication rule
    $$
    \text{joint} = \text{conditional}\times\text{marginal}
    $$
-   In many problems a joint distribution is nsturally described by specifying the marginal distribution of $X$ and the family of conditional distributions of $Y$ given values of $X$




```{r}
#| echo: false

N_rep = 16000

# first roll 
u1 = sample(1:4, size = N_rep, replace = TRUE)

# second roll
u2 = sample(1:4, size = N_rep, replace = TRUE)

# sum
x = u1 + u2

# max
y = pmax(u1, u2)
```


```{r}
#| echo: false

dice_sim = data.frame(1:N_rep, u1, u2, x, y)

# dice_sim |>
#   head() |>
#   kbl(col.names = c("Repetition", "First roll", "Second roll", "X (sum)", "Y (max)")) |>
#   kable_styling(fixed_thead = TRUE) |>
#     row_spec(which(head(y) == 4), bold = TRUE, color = "white", background = "#FFA500")

```



```{r}
#| label: dice-sim-tile
#| echo: false
#| fig-cap:
#|   - "Mosaic plot of conditional distributions of X given values of Y"
#|   - "Mosaic plot of conditional distributions of Y given values of X"
#| layout-ncol: 2
#| warning: false


ggplot(dice_sim) +
  ggmosaic::geom_mosaic(
    aes(x = product(x, y), fill = x),
    offset = 0
  ) +
  scale_fill_viridis_d(option = "D") +
  ggmosaic::theme_mosaic() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12),
    panel.grid = element_blank()
  ) +
  labs(
    x = "y",
    fill = "x",
    y = "P(X = x | Y = y)"
  )

ggplot(dice_sim) +
  ggmosaic::geom_mosaic(
    aes(x = product(y, x), fill = y),
    offset = 0
  ) +
  scale_fill_viridis_d(option = "D") +
  ggmosaic::theme_mosaic() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12),
    panel.grid = element_blank()
  ) +
  labs(
    x = "x",
    fill = "y",
    y = "P(Y = y | X = x)"
  )

```







-   Be sure to distinguish between joint, conditional, and marginal distributions.
-   The joint distribution of $X$ and $Y$ is a distribution on $(X, Y)$ pairs. A mathematical expression of a joint distribution is a function of both values of $X$ and values of $Y$.
-   The conditional distribution of $Y$ given $X=x$ is a distribution on $Y$ values (among $(X, Y)$ pairs with a fixed value of $X=x$). A mathematical expression of a conditional distribution will involve both $x$ and $y$, but $x$ is treated like a fixed constant and $y$ is treated as the variable. Note: the possible values of $Y$ might depend on the value of $x$.
-   The marginal distribution of $Y$ is a distribution on $Y$ values only, regardless of the value of $X$. A mathematical expression of a marginal distribution will have only values of the single variable in it; for example, an expression for the marginal distribution of $Y$ will only have $y$ in it (no $x$, not even in the possible values).
-   **Be careful** when conditioning with continuous random variables. Remember that the probability that a continuous random variable is equal to a particular value is 0; that is, for continuous $X$, $\text{P}(X=x)=0$. - Mathematically, when we condition on $\{X=x\}$ we are really conditioning on $\{|X-x|<\epsilon\}$ --- the event that the random variable $X$ is within $\epsilon$ of the value $x$ --- and seeing what happens in the idealized limit when $\epsilon\to0$.
-   Practically, $\epsilon$ represents our "close enough" degree of precision, e.g., $\epsilon=0.01$ if "within 0.01" is close enough.
-   When conditioning on a continuous random variable $X$ in a simulation, *never* condition on $\{X=x\}$; rather, condition on $\{|X-x|<\epsilon\}$ where $\epsilon$ represents the suitable degree of precision.

## Conditional Expected Value


::: {#exm-dice-ce}
Roll a fair four-sided die twice.  Let $X$ be the sum of the two rolls, and let $Y$ be the larger of the two rolls (or the common value if a tie).

:::

| $p_{X, Y}(x, y)$ |      |      |      |      |            |
|------------------|-----:|-----:|-----:|-----:|-----------:|
| $x$ \\ $y$       |    1 |    2 |    3 |    4 | $p_{X}(x)$ |
| 2                | 1/16 |    0 |    0 |    0 |       1/16 |
| 3                |    0 | 2/16 |    0 |    0 |       2/16 |
| 4                |    0 | 1/16 | 2/16 |    0 |       3/16 |
| 5                |    0 |    0 | 2/16 | 2/16 |       4/16 |
| 6                |    0 |    0 | 1/16 | 2/16 |       3/16 |
| 7                |    0 |    0 |    0 | 2/16 |       2/16 |
| 8                |    0 |    0 |    0 | 1/16 |       1/16 |
| $p_Y(y)$         | 1/16 | 3/16 | 5/16 | 7/16 |            |


1. Compute and interpret $\text{E}(Y)$. How could you find a simulation-based approximation?
\
\
\
1.  We have seen that the long run average value of $Y$ is 3.125. Would you expect the conditional long run average value of $Y$ given $X = 8$ to be greater than, less than, or equal to 3.125? Explain without doing any calculations. What about given $X = 3$?\
    \
    \
    \
    \
6.  How could you use simulation to approximate the conditional long run average value of $Y$ given $X = 6$?\
    \
    \
    \
    \
1. Compute and interpret $\text{E}(Y|X=6)$.
\
\
\
1. Find $\text{E}(Y|X=x)$ for each possible value of $x$ of $X$.
\
\
\
\
\
1. Compute and interpret $\text{E}(X|Y = 4)$. How could you find a simulation-based approximation?
\
\
\
1. Find $\text{E}(X|Y = y)$ for each possible value $y$ of $Y$.
\
\
\
\
\

-   The **conditional expected value** (a.k.a. *conditional expectation* a.k.a. *conditional mean*), of a random variable $Y$ given the event $\{X=x\}$, defined on a probability space with measure $\text{P}$, is a *number* denoted $\text{E}(Y|X=x)$ representing the probability-weighted average value of $Y$, where the weights are determined by the conditional distribution of $Y$ given $X=x$.
    \begin{align*}
	& \text{Discrete $X, Y$ with conditional pmf $p_{Y|X}$:} & \text{E}(Y|X=x) & = \sum_y y p_{Y|X}(y|x)\\
    \end{align*}
-   Remember, when conditioning on $X=x$, $x$ is treated as a fixed constant. The conditional expected value  $\text{E}(Y | X=x)$ is  a *number* representing the mean of the conditional distribution of $Y$ given $X=x$.
-   The conditional expected value $\text{E}(Y | X=x)$ is the long run average value of $Y$ over only those outcomes for which $X=x$.
-   To approximate $\text{E}(Y|X = x)$, simulate many $(X, Y)$ pairs, discard the pairs for which $X\neq x$, and average the $Y$ values for the pairs that remain. 
