# Expected Value

{{< include _r_setup.qmd >}}


-   The distribution of a random variable specifies the possible values and the probability of any event that involves the random variable.
-   The distribution of a random variable contains all the information about its long run behavior. It is also useful to summarize some key features of a distribution.
-   One summary characteristic of a distribution is the **long run average value** of the random variable.
-   We can approximate the long run average value by simulating many values of the random variable and computing the average (mean) in the usual way: sum the simulated values and divide by the number of simulated values.

::: {#exm-matching-ev}
Recall the matching problem with $n=4$.
@tbl-matching-sim-10-ev displays the value of $X$, the number of matches, for 10 simulated repetitions.
:::

```{r}
#| label: tbl-matching-sim-10-ev
#| echo: false
#| tbl-cap: "The value of $X$, the number of matches, for 10 simulated repetitions of the matching problem with $n=4$."

kbl(
  data.frame(1:10, c(0, 1, 0, 0, 2, 0, 1, 1, 4, 2)),
  col.names = c("Repetition", "X (number of matches)")
) |>
  kable_styling(fixed_thead = TRUE)
```

1.  How could you use the simulation results to approximate the long run average value of $X$? How could you get a better approximation of the *long run* average?\
    \
    \
    \
    \
2.  Rather than adding the 10 values and dividing by 10, how could you simplify the calculation in the previous part?\
    \
    \
    \
    \
3.  @tbl-matching-sim-ev below summarizes 24000 simulated values of $X$. Approximate the long run average value of $X$.\
    \
    \
    \
    \
4.  Recall the distribution of $X$ from @exm-matching-pmf. What would be the corresponding mathematical formula for the theoretical long run average value of $X$? This number is called the "expected value" of $X$.\
    \
    \
    \
    \
5.  Is the expected value the most likely value of $X$?\
    \
    \
    \
    \
6.  Is the expected value of $X$ the "value that we would expect" on a *single* repetition of the matching problem?\
    \
    \
    \
    \
7.  Explain in what sense the expected value is "expected".\
    \
    \
    \
    \

```{r}
#| label: tbl-matching-sim-ev
#| echo: false
#| tbl-cap: "Summary of 24000 simulated values of $X$, the number of matches, in the matching problem with $n=4$."


# One repetition of the number of matches, for a given n
simulate_number_matches = function(n) {
  # sample(1:n) puts the values 1:n in random order
  sum(sample(1:n) == 1:n)
}

# Many repetitions, for n = 4
N_rep = 24000
number_matches = replicate(N_rep, simulate_number_matches(4))

# Summarize the simulated values
table(number_matches) |>
  as.data.frame() |>
  adorn_totals("row") |>
  kbl(col.names = c("Number of Matches (X)",
                    "Frequency")) |>
  kable_styling(fixed_thead = TRUE)
```

```{r}
#| include: false
#| eval: false

sum(number_matches)

sum(number_matches) / N_rep

mean(number_matches)
```

-   The **expected value** (a.k.a. *expectation* a.k.a. *mean*), of a random variable $X$ is a number denoted $\text{E}(X)$ representing the probability-weighted average value of $X$.
-   The expected value of a discrete random variable with pmf $p_X$ is defined as $$
    \text{E}(X) = \sum_x x p_X(x) = \sum \text{value}\times\text{probability}
    $$
-   Note well that $\text{E}(X)$ represents *a single number*.
-   The expected value is the "balance point" (center of gravity) of a distribution.
-   The expected value of a random variable $X$ is defined by the probability-weighted average according to the underlying probability measure. But the expected value can also be interpreted as the **long-run average value**, and so can be approximated via simulation.
-   Read the symbol $\text{E}(\cdot)$ as
    -   Simulate lots of values of what's inside $(\cdot)$
    -   Compute the average. This is a "usual" average; just sum all the simulated values and divide by the number of simulated values.
    
```{r}
#| label: tbl-matching-lln
#| echo: false
#| tbl-cap: "Running average of 10 simulated values of $X$ in the matching problem with $n=4$."

x = c(0, 1, 0, 0, 2, 0, 1, 1, 4, 2)

n = length(x)

xbar = cumsum(x) / (1:n)

kbl(data.frame(1:n, x, xbar),
    align = "r",
    col.names = c("Repetition", "Value of X", "Running average of X"),
    digits = 3) |>
  kable_styling(fixed_thead = TRUE)
```


```{r}
#| label: fig-matching-lln
#| echo: false
#| fig-cap: "Running average of 10 simulated values of $X$ in the matching problem with $n=4$."

plot(1:n, xbar, type = "o",
     xlab = "Repetition",
     ylab = "Running average of X",
     ylim = c(0, 2),
     col = "blue",
     xaxt = "n")
axis(1, 1:n)
abline(h = 1, lty = 2, col = "orange")

```

```{r}
#| label: fig-matching-lln2
#| echo: false
#| fig-cap: "Running average of 100 simulated values of $X$ in the matching problem with $n=4$."

x = c(x, replicate(90, simulate_number_matches(4)))
n = length(x)

xbar = cumsum(x) / (1:n)


plot(1:n, xbar, type = "l",
     xlab = "Repetition",
     ylab = "Running average of X",
     ylim = c(0, 2),
     col = "blue",
     xaxt = "n")
axis(1, seq(0, 100, 10))
abline(h = 1, lty = 2, col = "orange")

```

```{r}
#| label: fig-matching-lln3
#| echo: false
#| fig-cap: "Running average of 1000 simulated values of $X$ in the matching problem with $n=4$."

x = c(x, replicate(900, simulate_number_matches(4)))
n = length(x)

xbar = cumsum(x) / (1:n)


plot(1:n, xbar, type = "l",
     xlab = "Repetition",
     ylab = "Running average of X",
     ylim = c(0, 2),
     col = "blue",
     xaxt = "n")
axis(1, seq(0, 1000, 100))
abline(h = 1, lty = 2, col = "orange")

```


::: {#exm-binomial-stock-ev}
Continuing @exm-binomial-stock.
Consider an extremely simplified model for the daily closing price of a certain stock.
Every day the price either goes up or goes down, and the movements are independent from day-to-day.
Assume that the probability that the stock price goes up on any single day is 0.25.
Let $X$ be the number of days in which the price goes up in the next 5 days.
:::

1.  Suggest a shortcut formula for $\text{E}(X)$.\
    \
    \
    \
    \
2.  Compute $\text{E}(X)$ using the distribution of $X$ from @exm-binomial-stock. Did the shortcut formula work?\
    \
    \
    \
    \
3.  Interpret $\text{E}(X)$.\
    \
    \
    \
    \

-   If $X$ has a Binomial($n$, $p$) distribution then $\text{E}(X) = np$.

::: {#exm-poisson-hr-ev}
Continuing @exm-poisson-hr.
Assume that $X$, the number of home runs hit (in total by both teams) in a randomly selected Major League Baseball game, follows a Poisson(2.3) distribution.
:::

1.  Compute $\text{E}(X)$.\
    \
    \
    \
    \
    \
2.  Interpret $\text{E}(X)$ in context.\
    \
    \
    \
    \

-   If $X$ has a Poisson($\mu$) distribution then $\text{E}(X) = \mu$.
