# Homework 7 solutions {.unnumbered}


```{r}
#| echo: false
#| warning: false
#| message: false

library(knitr)
library(kableExtra)
library(ggplot2)
library(janitor)
```

## Problem 1

Assume a Poisson($\mu$) model for the number of home runs hit (in total by both teams) in a MLB game.
Let $X_1, \ldots, X_n$ be a random sample of home run counts for  $n$ games.

Suppose we want to estimate $\theta = \mu e^{-\mu}$, the probability that any single game has exactly 1 HR. Consider two estimators of $\theta$

- $\hat{\theta} = \bar{X} e^{-\bar{X}}$
- $\hat{p} =\text{sample proportion of 1s} = \frac{\text{number of games in the sample with 1 HR}}{\text{sample size}}$

a.  Compute the value of $\hat{\theta}$ based on the sample (3, 0, 1, 4, 0). Write a clearly worded sentence reporting in context this estimate of $\theta$.
a.  Compute the value of $\hat{p}$ based on the sample (3, 0, 1, 4, 0). Write a clearly worded sentence reporting in context your estimate of $\theta$.
a.  Which of these two estimators is the MLE of $\theta$ in this situation? Explain, without doing any calculations.
a.  It can be shown that $\hat{p}$ is an unbiased estimator of $\theta$. Explain what this means.
a.  Is $\hat{\theta}$ an unbiased estimator of $\theta$? Explain. (You don't have to derive anything; just apply a general principle.)
a.  Suppose $\mu = 2.3$ and $n=5$. Explain in full detail how you would use simulation to approximate the bias of $\hat{\theta}$.
a.  Conduct the simulation from the previous part and plot the approximate of $\hat{\theta}$ when $\mu = 2.3$ and $n = 5$.
a.  Explain in full detail how you would use simulation to approximate the bias function of $\hat{\theta}$ when $n=5$.
a.  Conduct the simulation from the previous part and plot the approximate bias function when $n=5$. For what values of $\mu$ does $\hat{\theta}$ tend to overestimate $\mu$? Underestimate? For what values of $\mu$ is the bias the worst?




### Solution


a.  For the sample (3, 0, 1, 4, 0), the value of $\bar{X}$ is 8/5 and  the value of $\hat{\theta}$ is $(8/5)e^{-8/5}=0.323$. Using this estimator, based on this sample we would estimate that 32.3 percent of baseball games have 1 home run.
a.  For the sample (3, 0, 1, 4, 0) there is 1 game with 1 HR so $\hat{p} = 1/5 = 0.2$. Using this estimator, based on this sample we would estimate that 20 percent of baseball games have 1 home run.
a.  We know that the MLE of $\mu$ for a Poisson distribution is $\bar{X}$, so by the invariance principle of MLEs, $\bar{X}e^{-\bar{X}}$ is the MLE of $\mu e^{-\mu}$.
a.  Over many samples, $\hat{p}$ does not consistent over or under estimate $\theta$.
a.  No. $\bar{X}$ is an unbiased estimator of $\mu$ but nonlinear transformations do not preserve unbiasedness.
$\hat{\theta}$ is not unbiased estimator of $\theta$.
    $$
    \text{E}(\hat{\theta}) = \text{E}(\bar{X}e^{-\bar{X}})\neq     \text{E}(\bar{X})e^{-\text{E}(\bar{X})} = \mu e^{-\mu} = \theta
    $$
a.  For $\mu = 2.3$
    -   Simulate a sample of size $n$ from a Poisson(2.3) distribution, say (3, 0, 1, 4, 0) for $n=5$.
    -   For the simulated sample compute the sample mean $\bar{X}$ and the value of $\hat{\theta} = \bar{X}e^{-\bar{X}}$, say 8/5 and $(8/5)e^{-8/5}=0.323$.
    -   Repeat many times, simulating many samples of size $n$ and many values of $\hat{\theta}$ for the chosen $\mu$.
    -   Average the values of $\hat{\theta}$ and subtract $\theta=2.3 e^{-2.3}$ to approximate the bias when $\mu = 2.3$.
a.  See below.
a.  We carry out the process in the previous part for many values of $\mu$. Choose a grid of $\mu$ values, say 0.01, 0.02, $\ldots$, 10
    -   Choose a value of $\mu$ from the grid, say 2.3
    -   Simulate a sample of size $n$ from a Poisson($\mu$) distribution, say (3, 0, 1, 4, 0) for $n=5$.
    -   For the simulated sample compute the sample mean $\bar{X}$ and the value of $\hat{\theta} = \bar{X}e^{-\bar{X}}$, say 8/5 and $(8/5)e^{-8/5}=0.323$.
    -   Repeat many times, simulating many samples of size $n$ and many values of $\hat{\theta}$ for the chosen $\mu$.
    -   Average the values of $\hat{\theta}$ and subtract $\theta=\mu e^{-\mu}$ to approximate the bias for the chosen $\mu$.
    -   Repeat the above process for all the values of $\mu$ in the grid and plot the approximate bias as a function of $\mu$ to approximate the bias function.
a.  See below. 




## Problem 2

Continuing Problem 2.

a.  It can be shown that $\text{Var}(\hat{p}) = \frac{\theta(1-\theta)}{n}$. Compute  $\text{Var}(\hat{p})$ when $\mu = 2.3$ and $n=5$. Then write a clearly worded sentence interpreting this value.
a.  Suppose $\mu = 2.3$ and $n=5$. Explain in full detail how you would use simulation to approximate the variance of $\hat{\theta}$.
a.  Conduct the simulation from the previous part and approximate  the variance of $\hat{\theta}$ when $\mu = 2.3$ and $n=5$. Then write a clearly worded sentence interpreting this value.
a.  Which estimator has smaller variance when $\mu = 2.3$ (and $n=5$)? Answer, but then explain why this information alone is not really useful.
a.  Explain in full detail how you would use simulation to approximate the variance function of $\hat{\theta}$ (if $n=5$).
a.  Conduct the simulation from the previous part and plot the approximate variance function. Compare to the variance function of $\hat{p}$. Based on variability alone, which estimator is preferred?



### Solution

a.  When $\mu = 2.3$, $\theta = 2.3e^{-2.3} = 0.231$, so if $n=5$
    $$
    \text{Var}_{\mu = 2.3}(\hat{p}) = \frac{0.231(1-0.231)}{5} = 0.035
    $$ 
    This number measures the sample-to-sample variance of sample proportions of games with 1 HR over many samples of 5 games, assuming that $\mu = 2.3$.
a.  For $\mu = 2.3$
    -   Simulate a sample of size $n$ from a Poisson(2.3) distribution, say (3, 0, 1, 4, 0) for $n=5$.
    -   For the simulated sample compute the sample mean $\bar{X}$ and the value of $\hat{\theta} = \bar{X}e^{-\bar{X}}$, say 8/5 and $(8/5)e^{-8/5}=0.323$.
    -   Repeat many times, simulating many samples of size $n$ and many values of $\hat{\theta}$ for the chosen $\mu$.
    -   Average the values of $\hat{\theta}$, for each value of $\hat{\theta}$ find its deviation from the average, and then average the squared deviations. This number approximates $\text{Var}(\hat{\theta})$ when $\mu = 2.3$
a.  See below. The value measures the sample-to-sample variance of values of $\hat{\theta}$ (estimates of the probability that a game has 1 HR), assuming that $\mu = 2.3$.
a.  $\hat{\theta}$ has smaller variance when $\mu = 2.3$, but that by itself isn't useful because we don't know what $\mu$ is.
a.  We carry out the process in the previous part for many values of $\mu$. Choose a grid of $\mu$ values, say 0.01, 0.02, $\ldots$, 10
    -   Choose a value of $\mu$ from the grid, say 2.3
    -   Simulate a sample of size $n$ from a Poisson($\mu$) distribution, say (3, 0, 1, 4, 0) for $n=5$.
    -   For the simulated sample compute the sample mean $\bar{X}$ and the value of $\hat{\theta} = \bar{X}e^{-\bar{X}}$, say 8/5 and $(8/5)e^{-8/5}=0.323$.
    -   Repeat many times, simulating many samples of size $n$ and many values of $\hat{\theta}$ for the chosen $\mu$.
    -   Average the values of $\hat{\theta}$, for each value of $\hat{\theta}$ find its deviation from the average, and then average the squared deviations. This number approximates $\text{Var}(\hat{\theta})$ for the chosen $\mu$
    -   Repeat the above process for all the values of $\mu$ in the grid and plot the approximate bias as a function of $\mu$ to approximate the bias function.
a.  See below. Seems like $\hat{\theta}$ has smaller variance regardless of $\mu$.




## Problem 3

Continuing Problems 2 and 3

a.  Compute $\text{MSE}(\hat{p})$ when $\mu = 2.3$ and $n=5$. (You can do the next part first if you want, but it helps to work with specific numbers first.)
a.  Derive the MSE function of $\hat{p}$. (Hint: see facts from previous parts.)
a.  Suppose $\mu = 2.3$ (and $n=5$). Explain in full detail how you would use simulation to approximate the MSE of $\hat{\theta}$.
a.  Conduct the simulation from the previous part and approximate the MSE of $\hat{\theta}$ when $\mu =2.3$ (and $n=5$).
a.  Which estimator has smaller MSE when $\mu = 2.3$ (and $n=5$)? Answer, but then explain why this information alone is not really useful.
a.  Conduct the simulation from the previous part and plot the approximate MSE function. Compare to the MSE function of $\hat{p}$.
a.  Compare the MSEs of the two estimators for $n=5$ and a few other values of $n$. Is there a clear preference between these two estimators? Discuss.



### Solution


a.  Since $\hat{p}$ is unbiased, its MSE is just its variance.
a.  Since $\hat{p}$ is unbiased, its MSE function is just its variance function.
a.  For $\mu = 2.3$
    -   Simulate a sample of size $n$ from a Poisson(2.3) distribution, say (3, 0, 1, 4, 0) for $n=5$.
    -   For the simulated sample compute the sample mean $\bar{X}$ and the value of $\hat{\theta} = \bar{X}e^{-\bar{X}}$, say 8/5 and $(8/5)e^{-8/5}=0.323$.
    -   Repeat many times, simulating many samples of size $n$ and many values of $\hat{\theta}$ for the chosen $\mu$.
    -   For each value of $\hat{\theta}$ subtract $\theta=2.3 e^{-2.3}$ and square the difference, then average these squared differences to approximate the MSE when $\mu = 2.3$.
a.  See below. The squared bias is fairly small compared to the variance, so the MSE is pretty close to the variance function. 
a.  $\hat{\theta}$ has smaller MSE when $\mu = 2.3$, but this isn't really useful because we don't know what $\mu$ is.
a.  See below.
a.  It seems that (at least for the values of $n$ that we considered) $\hat{\theta}$ has smaller MSE than $\hat{p}$ for all $\mu$.
The sample proportion doesn't use the fact that the population distribution is Poisson, and it treats each value as either 1 or not. $\bar{X} e^{-\bar{X}}$ uses all the data and the fact that the population distribution is Poisson, so its better tuned to this problem.


## Simulation

### For mu = 2.3

```{r}
n = 5

mu = 2.3

N_samples = 10000

theta_hats = vector(length = N_samples)

for (i in 1:N_samples) {
  # simulate a sample of size n from the Poisson(mu) population
  x = rpois(n, mu)
  
  # compute theta-hat
  theta_hats[i] = mean(x) * exp(-mean(x))
}

# Approximate the bias of theta-hat for this mu
mean(theta_hats) - mu * exp(-mu)

# Approximate the variance of theta-hat for this mu
var(theta_hats)

# Approximate the MSE of theta-hat for this mu
mean((theta_hats - mu * exp(-mu)) ^ 2)

```


### For all mu

```{r}
n = 5

# grid of mu values: 0.1, 0.2, 0.3, ..., 4.9, 5
mus = seq(0.1, 5, 0.1)

N_samples = 10000

# initialize vectors that will record bias, var, mse as functions of mu
bias_thetahat = vector()
var_thetahat = vector()
mse_thetahat = vector()

for (mu in mus) {
  
  # true value of theta for this mu
  theta = mu * exp(-mu)
  
  # for each mu simulate many samples and many values of sigma-hat^2
  theta_hats = vector(length = N_samples)
  
  for (i in 1:N_samples) {
    # simulate a sample of size n from the Poisson(mu) population
    x = rpois(n, mu)
    
    # compute theta-hat
    theta_hats[i] = mean(x) * exp(-mean(x))
    
  }
  
  # Approximate bias of theta-hat for this mu and add to the list
  bias_thetahat = c(bias_thetahat, mean(theta_hats) - theta)
  
  # Approximate variance of sigma-hat^2 for this mu and add to the list
  var_thetahat = c(var_thetahat, var(theta_hats))
  
  # Approximate MSE of sigma-hat^2 for this mu and add to the list
  # Can just do bias^2 + var, but can also use definition of MSE
  mse_thetahat = c(mse_thetahat, mean((theta_hats - theta) ^ 2))
  
}

```



### Bias function

Any "wiggles" are due to natural simulation margin of error (can make these go away by simulating even more samples)

```{r}
plot(mus, bias_thetahat,
     xlab = "mu",
     ylab = "Bias of thetahat")

```

### Squared bias function

Any "wiggles" are due to natural simulation margin of error (can make these go away by simulating even more samples)

```{r}
plot(mus, bias_thetahat ^ 2,
     xlab = "mu",
     ylab = "Squared Bias of thetahat",
     ylim = c(0, 0.05)) # set common y-axis scale

```

### Variance function

Any "wiggles" are due to natural simulation margin of error (can make these go away by simulating even more samples)

```{r}
plot(mus, var_thetahat,
     xlab = "mu",
     ylab = "Variance",
     ylim = c(0, 0.05), # set common y-axis scale
     main = "Variance functions of theta-hat (dots) and p-hat (dashed)")
curve(x * exp(-x) * (1 - x * exp(-x)) / n, add = TRUE,
      lty = 2)

```


### MSE function

Any "wiggles" are due to natural simulation margin of error (can make these go away by simulating even more samples)

```{r}
plot(mus, mse_thetahat,
     xlab = "mu",
     ylab = "MSE",
     ylim = c(0, 0.05), # set common y-axis scale
     main = "MSE functions of theta-hat (dots) and p-hat (dashed)")
curve(x * exp(-x) * (1 - x * exp(-x)) / n, add = TRUE,
      lty = 2)

```

### Different n

```{r}
n = 50

# grid of mu values: 0.1, 0.2, 0.3, ..., 4.9, 5
mus = seq(0.1, 5, 0.1)

N_samples = 10000

# initialize vectors that will record bias, var, mse as functions of mu
bias_thetahat = vector()
var_thetahat = vector()
mse_thetahat = vector()

for (mu in mus) {
  
  # true value of theta for this mu
  theta = mu * exp(-mu)
  
  # for each mu simulate many samples and many values of sigma-hat^2
  theta_hats = vector(length = N_samples)
  
  for (i in 1:N_samples) {
    # simulate a sample of size n from the Poisson(mu) population
    x = rpois(n, mu)
    
    # compute theta-hat
    theta_hats[i] = mean(x) * exp(-mean(x))
    
  }
  
  # Approximate bias of theta-hat for this mu and add to the list
  bias_thetahat = c(bias_thetahat, mean(theta_hats) - theta)
  
  # Approximate variance of sigma-hat^2 for this mu and add to the list
  var_thetahat = c(var_thetahat, var(theta_hats))
  
  # Approximate MSE of sigma-hat^2 for this mu and add to the list
  # Can just do bias^2 + var, but can also use definition of MSE
  mse_thetahat = c(mse_thetahat, mean((theta_hats - theta) ^ 2))
  
}

```


```{r}
plot(mus, mse_thetahat,
     xlab = "mu",
     ylab = "MSE",
     ylim = c(0, 0.006), # set common y-axis scale
     main = "n = 50:MSE functions of theta-hat (dots) and p-hat (dashed)")
curve(x * exp(-x) * (1 - x * exp(-x)) / n, add = TRUE,
      lty = 2)
```



```{r}
n = 20

# grid of mu values: 0.1, 0.2, 0.3, ..., 4.9, 5
mus = seq(0.1, 5, 0.1)

N_samples = 10000

# initialize vectors that will record bias, var, mse as functions of mu
bias_thetahat = vector()
var_thetahat = vector()
mse_thetahat = vector()

for (mu in mus) {
  
  # true value of theta for this mu
  theta = mu * exp(-mu)
  
  # for each mu simulate many samples and many values of sigma-hat^2
  theta_hats = vector(length = N_samples)
  
  for (i in 1:N_samples) {
    # simulate a sample of size n from the Poisson(mu) population
    x = rpois(n, mu)
    
    # compute theta-hat
    theta_hats[i] = mean(x) * exp(-mean(x))
    
  }
  
  # Approximate bias of theta-hat for this mu and add to the list
  bias_thetahat = c(bias_thetahat, mean(theta_hats) - theta)
  
  # Approximate variance of sigma-hat^2 for this mu and add to the list
  var_thetahat = c(var_thetahat, var(theta_hats))
  
  # Approximate MSE of sigma-hat^2 for this mu and add to the list
  # Can just do bias^2 + var, but can also use definition of MSE
  mse_thetahat = c(mse_thetahat, mean((theta_hats - theta) ^ 2))
  
}

```


```{r}
plot(mus, mse_thetahat,
     xlab = "mu",
     ylab = "MSE",
     ylim = c(0, 0.015), # set common y-axis scale
     main = "n = 20:MSE functions of theta-hat (dots) and p-hat (dashed)")
curve(x * exp(-x) * (1 - x * exp(-x)) / n, add = TRUE,
      lty = 2)
```



Xavier and Yolanda are playing roulette. They both place bets on red on the same 3 spins of the roulette wheel before Xavier has to leave. (Remember, the probability that any bet on red on a single spin wins is 18/38.) After Xavier leaves, Yolanda places bets on red on 2 more spins of the wheel. Let $X$ be the number of bets that Xavier wins and  let $Y$ be the number that Yolanda wins.

a. Identify by name the marginal distribution of $X$. Be sure to specify the values of any relevant parameters. Compute $\text{E}(X)$.
a. Identify by name the marginal distribution of $Y$. Be sure to specify the values of any relevant parameters. Compute $\text{E}(Y)$.
a. The joint distribution of $X$ and $Y$ is represented in the table below. Explain why $p_{X, Y}(1, 4) = 0$ and $p_{X, Y}(2, 1) = 0$.
a. Compute $p_{X, Y}(2, 3)$. (Yes, the table tells you it's 0.1766, but you have to show how this number can be computed based on the assumptions of the problem.)
a. Are $X$ and $Y$ independent? Justify your answer with an appropriate calculation.
a. Compute and interpret $\text{P}(X + Y) = 4$.
a. Make a table representing the marginal distribution of $X+Y$.
a. Use the table from the previous part to compute $\text{E}(X+Y)$, and verify that it is equal to $\text{E}(X)+\text{E}(Y)$.
a. Without doing any computation, determine if $\text{Var}(X+Y)$ will be greater than, less than, or equal to $\text{Var}(X) + \text{Var}(Y)$. Explain your reasoning.


| $x$, $y$ |      0 |      1 |      2 |      3 |      4 |      5 |
|----------|-------:|-------:|-------:|-------:|-------:|-------:|
| 0        | 0.0404 | 0.0727 | 0.0327 |      0 |      0 |      0 |
| 1        |      0 |  0.109 | 0.1963 | 0.0883 |      0 |      0 |
| 2        |      0 |      0 | 0.0981 | 0.1766 | 0.0795 |      0 |
| 3        |      0 |      0 |      0 | 0.0294 |  0.053 | 0.0238 |


### Solution

a. $X$ has a Binomial distribution with parameters $n=3$ and $p = 18/38$, so $\text{E}(X) = np = 3(18/38) = 1.42$.
a. $Y$ has a Binomial distribution with parameters $n=5$ and $p = 18/38$, so $\text{E}(X) = np = 5(18/38) = 2.37$.
a. In the first 3 bets Yolanda wins the same number of games as Xavier, and Yolanda only plays two more games. So $Y$ can't be more than $X+2$, and $Y$ can't be less than $X$.
a. To have $X = 2$ and $Y=3$ we need to have exactly two wins in the first 3 bets (where both play) exactly 1 win in the last 2 bets (Yolanda only). The number of wins in the first 3 bests is Binomial(3, 18/38), the number of wins in the last two bets is Binomial(2, 18/38) and the first 3 bets are independent of the last two bets. So the probability is `dbinom(2, 3, 18 / 38) * dbinom(1, 2, 18 / 38)` = `r round(dbinom(2, 3, 18 / 38) * dbinom(1, 2, 18 / 38), 3)` or
    $$
    p_{X, Y}(2, 3) = \text{P}(X = 2, Y = 3) = \left(\binom{3}{1}(18/38)^2(20/38)^1\right)\left(\binom{2}{1}(18/38)^1(20/38)^1\right) = 0.177
    $$
a. No. $X$ and $Y$ are not independent. For example $\text{P}(X = 0) = 0.15$ but $\text{P}(X=0|Y = 0) = 1$.
a. Sum over the $(X, Y)$ pairs for which $X + Y = 4$: (0, 4), (1, 3), (2, 2), (3, 1), (4, 0). So 0 + 0.0883 + 0.0981 + 0 = 0.1864. Suppose they do this every time they visit the casino. Over many visits to the casino, together they win a total of 4 bets in 18.6% of visits
a. The possible values are $0, 1, \ldots, 8$. Fill in the probabilities like in the previous part. (Alternative $X + Y$ has the same distribution as $2X + Z$ where $X$ and $Z$ are independent and $Z$ (number of wins in last two bets) has a Binomial(2, 18/38) distribution).

    | Value | Probability |
    |-------|------------:|
    | 0     |      0.0404 |
    | 1     |      0.0727 |
    | 2     |      0.1417 |
    | 3     |      0.1963 |
    | 4     |      0.1864 |
    | 5     |      0.1766 |
    | 6     |      0.1089 |
    | 7     |      0.0530 |
    | 8     |      0.0238 |
     
a. From the table $\text{E}(X+Y) = 0(0.404) + 1(0.0727 + \cdots + 8(0.0238) = 3.79$, and  $\text{E}(X)+\text{E}(Y) = 3(18/38) + 5(18/38) = 3.79$. (These are exactly equal; any differences are just rounding errors)
a. There is a positive association so $\text{Var}(X+Y)$ will be greater than $\text{Var}(X) + \text{Var}(Y)$.




## Problem 3


Percent returns for assets $X$, $Y$, and $Z$ follow a joint distribution with

- Mean 10 and standard deviation 15 for asset X,
- Mean 5 and standard deviation 3 for asset Y,
- Correlation of −0.6 between asset X and asset Y
- Asset Z yields a constant return of 1 percent.

An investment portfolio has 60% of its funds in asset X, 30% in asset Y, and 10% in asset Z.

a. Let $R$ be the portfolio return. Express $R$ in terms of $X, Y, Z$. 
a. Compute $\text{E}(R)$.
a. Compute $\text{Cov}(X, Y)$.
a. Compute $\text{Cov}(X, Z)$.
a. Compute $\text{SD}(R)$.


### Solution



a. $R = 0.6X + 0.3Y + 0.1Z$
a. Linearity of expected value
    $$
    \text{E}(R)=\text{E}(0.6X + 0.3Y + 0.1Z) = 0.6\text{E}(X) + 0.3\text{E}(Y) + 0.1\text{E}(Z) = 0.6(10) + 0.3(5) + 0.1(1) = 7.6
    $$
a. $\text{Cov}(X, Y) = \text{Corr}(X, Y)\text{SD}(X)\text{SD}(Y) = (-0.6)(15)(3) = -27$
a. $Z$ is constant so $\text{Cov}(X, Z)=0$
a. Use properties of variance and covariance
\begin{align*}
& \text{Var}(0.6X + 0.3Y + 0.1Z) & &  
\\ & = \text{Var}(0.6X + 0.3Y) & & \text{$Z$ is constant, no affect variability}\\
 & = \text{Var}(0.6X) + \text{Var}(0.3Y) + 2\text{Cov}(0.6X, 0.3Y) & & \text{variance of sum}\\
 & = 0.6^2\text{Var}(X) + 0.3^2\text{Var}(Y) + 2(0.6)(0.3)\text{Cov}(X, Y) && \text{rules for variance}\\
 & =  0.6^2\text{SD}(X)^2 + 0.3^2\text{SD}(Y)^2 + 2(0.6)(0.3)\text{Corr}(X, Y)\text{SD}(X)\text{SD}(Y) && \text{correlation}\\
 & = 0.6^2(15^2) + 0.3^2(3)^2 + 2(0.6)(0.3)(-0.6)(15)(3) & & \\
 & = 72.09
 \end{align*}
 So $\text{SD}(R) = \sqrt{\text{Var}(R)} = \sqrt{72} = 8.49$.



 
## Problem 4

Devi and Paxton are meeting.
Arrival times are measured in minutes after noon, with negative times representing arrivals before noon.
Devi's arrival time follows a Normal distribution with mean 20 and SD 15 minutes, and Paxton's arrival time follows a Normal distribution with mean 25 and SD 10 minutes, independently of each other.

For each of the following, find the appropriate standardized value and make a ballpark estimate of the probability. Then use software to compute the probability.

a. Compute the probability that Devi arrives before noon.
a. Compute the probability that Devi arrives first.
a. Compute the probability that the first person to arrive has to wait more than 15 minutes for the second person to arrive.
a. **Coding required.** Code and run a simulation and use the results to approximate the probability from the previous part (just part c). You should first simulate (Devi, Paxton) arrival time pairs, and go from there.

### Solution


a. If $V$ is Devi's arrival time, $V\sim N(20, 15)$. We want $\text{P}(V < 0)$. The standardized value for 0 is $(0 - 20)/15=-1.333$; that is, 0 is 1.333 SDs below the mean. From the empirical rule, 87% of values are within 1.5 SDs of the mean, so our estimated probability is more than 6.5%. For software, see below; $\text{P}(V < 0)=0.09$.
a. If $X$ is Paxton's arrival time, we want $\text{P}(X > V) = \text{P}(X - V >0)$. Since $X$ and $V$ are independent each with a Normal distribution, then $X - V$ has a Normal distribution. The mean is $20-15 = 5$, and since $X$ and $V$ are independent
    $$
    \text{Var}(X - V) = \text{Var}(X) + \text{Var}(V) - 2\text{Cov}(X, V) = 10^2 + 15^2 - 0 = 325
    $$
    So $\text{SD}(X - V) = \sqrt{325} = 18.0$. 
    We want $\text{P}(X - V > 0)$. The standardized value for 0 is $(0 - 5)/18=-0.277$; that is, 0 is 0.277 SDs below the mean.
    The probability will be more than 50%. For software, see below; $\text{P}(X - V > 0)=0.61$.
a. We want $\text{P}(|X - V| >15) = \text{P}(X - V > 15) + \text{P}(X - V < -15)$. The standardized value for 15 is $(15 - 5) / 18 = 0.556$. The standardized value for -15 is $(-15 - 5) / 18 = -1.11$. The probability that a value is more than 1 SD below the mean is about 0.16; the probability that a value is more than 0.5 SDs above the mean is about 0.31. So the probability that we want is less than 0.47. It's $\text{P}(|X - V| > 15) = 0.42$.


```{r}
pnorm((0 - 20) / 15)

pnorm(0, 20, 15)
```

```{r}
1 - pnorm((0 - 5) / 18)

1 - pnorm(0, 5, 18)
```


```{r}
1 - pnorm((15 - 5) / 18) + pnorm((-15 - 5) / 18)

1 - pnorm(15, 5, 18) + pnorm(-15, 5, 18)
```





```{r}
N_rep <- 10000

V <- rnorm(N_rep, 20, 15)
X <- rnorm(N_rep, 25, 10)

sum(abs(V - X) > 15) / N_rep

cor(V, X)

mean(V)

sd(V)

mean(X)

sd(X)
```


```{r}
plot(V, X, col = ifelse(abs(V - X) > 15, "orange", "gray"))
abline(a = 15, b = 1, col = "orange")
abline(a = -15, b = 1, col = "orange")
```



```{r}
ggplot(data.frame(V, X), aes(x = V, y = X)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  geom_abline(slope = 1, intercept = c(-15, 15), col = "orange", size = 2)
```

## Problem 5 part 1

(Continued.) Devi and Paxton are meeting. Arrival times are measured in minutes after noon, with negative times representing arrivals before noon. Devi’s arrival time follows a Normal distribution with mean 20 and SD 15 minutes, and Paxton’s arrival time follows a Normal distribution with mean 25 and SD 10 minutes.

For each of the following, find the appropriate standardized value and make a ballpark estimate of the probability. Then use software to compute the probability.

Assume the pairs of arrival times follow a Bivariate Normal distribution with correlation 0.8

a. Compute the probability that Devi arrives first given that Paxton arrives at 12:10.
a. Compute the probability that the first person to arrive has to wait more than 15 minutes for the second person to arrive.
a. **Coding required.** Code and run a simulation and use the results to approximate the probability from the previous part (just part b). You should first simulate (Devi, Paxton) arrival time pairs, and go from there.


### Solution



Let $V$ be Devi's arrival time and let $X$ be Paxton's arrival time.


a. We want $\text{P}(V < X | X = 10) = \text{P}(V < 10 | X = 10)$. Paxton's time is 1.5 SDs below the mean, $\frac{10 - 25}{10} = -1.5$. Devi's conditional mean arrival time is 1.2 SDs below the mean, $0.8(-1.5) = -1.2$, so Devi's conditional mean arrival time is $\text{E}(V | X = 10) = 20 -1.2(15) =2$. Devi's conditional SD is $\text{SD}(V|X = 10) = 15\sqrt{1-0.8^2} = 9$. So $(V | (X = 10))\sim N(2, 9)$. $\text{P}(V < 10 | X = 10) = \text{P}(Z < \frac{10-2}{9}) = \text{P}(Z < 0.889)$; ballpark estimate a little less than 84%. Software (below) gives 0.81.
a. We want $\text{P}(|X - V| >15) = \text{P}(X - V > 15) + \text{P}(X - V < -15)$.  Since $X$ and $V$ have a Bivariate Normal distribution, then $X - V$ has a Normal distribution. The mean is $20-15 = 5$, and
    $$
    \text{Var}(X - V) = \text{Var}(X) + \text{Var}(V) - 2\text{Cov}(X, V) = 10^2 + 15^2 - 2(0.8)(10)(15) = 85
    $$
    So $\text{SD}(X - V) = \sqrt{85} = 9.2$.
    The standardized value for 15 is $(15 - 5) / 9.2 = 1.08$. The standardized value for -15 is $(-15 - 5) / 9.2 = -2.17$. The probability that a value is more than 2 SDs below the mean is about 0.025; the probability that a value is more than 1 SD above the mean is about 0.16. So the probability that we want is less than 0.185. It's $\text{P}(|X - V| > 15) = 0.15$.



```{r}
pnorm((10 - 2) / 9)

pnorm(10, 2, 9)
```

```{r}
1 - pnorm((15 - 5) / 9.2) + pnorm((-15 - 5) / 9.2)

1 - pnorm(15, 5, 9.2) + pnorm(-15, 5, 9.2)
```





```{r}
N_rep <- 10000

V <- rnorm(N_rep, 20, 15)
X <- rnorm(N_rep, 25 + 10 * 0.8 * (V - 20) / 15, 10 * sqrt(1 - 0.8^2))

sum(abs(V - X) > 15) / N_rep

cor(V, X)

mean(V)

sd(V)

mean(X)

sd(X)
```


```{r}
plot(V, X, col = ifelse(abs(V - X) > 15, "orange", "gray"))
abline(a = 15, b = 1, col = "orange")
abline(a = -15, b = 1, col = "orange")
```



```{r}
ggplot(data.frame(V, X), aes(x = V, y = X)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  geom_abline(slope = 1, intercept = c(-15, 15), col = "orange", size = 2)
```





## Problem 5 part 2

(Continued.) Devi and Paxton are meeting. Arrival times are measured in minutes after noon, with negative times representing arrivals before noon. Devi’s arrival time follows a Normal distribution with mean 20 and SD 15 minutes, and Paxton’s arrival time follows a Normal distribution with mean 25 and SD 10 minutes.

For each of the following, find the appropriate standardized value and make a ballpark estimate of the probability. Then use software to compute the probability.

Assume the pairs of arrival times follow a Bivariate Normal distribution with correlation -0.7

a. Compute the probability that Devi arrives first given that Paxton arrives at 12:10.
a. Compute the probability that the first person to arrive has to wait more than 15 minutes for the second person to arrive.
a. **Coding required.** Code and run a simulation and use the results to approximate the probability from the previous part (just part b). You should first simulate (Devi, Paxton) arrival time pairs, and go from there.


### Solution

a. We want $\text{P}(V < X | X = 10) = \text{P}(V < 10 | X = 10)$. Paxton's time is 1.5 SDs below the mean, $\frac{10 - 25}{10} = -1.5$. Devi's conditional mean arrival time is 1.05 SDs *above* the mean, $(-0.7)(-1.5) = 1.05$, so Devi's conditional mean arrival time is $\text{E}(V | X = 10) = 20 +1.05(15) =35.75$. Devi's conditional SD is $\text{SD}(V|X = 10) = 15\sqrt{1-0.7^2} = 10.7$. So $(V | (X = 10))\sim N(35.75, 10.7)$. $\text{P}(V < 10 | X = 10) = \text{P}(Z < \frac{10-35.75}{10.7}) = \text{P}(Z < -2.4)$; ballpark estimate is about 0.005. Software (below) gives 0.008.
a. We want $\text{P}(|X - V| >15) = \text{P}(X - V > 15) + \text{P}(X - V < -15)$.  Since $X$ and $V$ have a Bivariate Normal distribution, then $X - V$ has a Normal distribution. The mean is $20-15 = 5$, and
    $$
    \text{Var}(X - V) = \text{Var}(X) + \text{Var}(V) - 2\text{Cov}(X, V) = 10^2 + 15^2 - 2(-0.7)(10)(15) = 535
    $$
    So $\text{SD}(X - V) = \sqrt{535} = 23.1$.
    The standardized value for 15 is $(15 - 5) / 23.1 = 0.43$. The standardized value for -15 is $(-15 - 5) / 23.1 = -0.86$. The probability that a value is more than 1 SD below the mean is about 0.16; the probability that a value is more than 0.5 SD above the mean is about 0.31. So the probability that we want is about 0.47. It's $\text{P}(|X - V| > 15) = 0.53$.





```{r}
pnorm((10 - 35.75) / 10.7)

pnorm(10, 35.75, 10.7)
```

```{r}
1 - pnorm((15 - 5) / 23.1) + pnorm((-15 - 5) / 23.1)

1 - pnorm(15, 5, 23.1) + pnorm(-15, 5, 23.1)
```





```{r}
N_rep <- 10000

V <- rnorm(N_rep, 20, 15)
X <- rnorm(N_rep, 25 + 10 * (-0.7) * (V - 20) / 15, 10 * sqrt(1 - 0.7^2))

sum(abs(V - X) > 15) / N_rep

cor(V, X)

mean(V)

sd(V)

mean(X)

sd(X)
```


```{r}
plot(V, X, col = ifelse(abs(V - X) > 15, "orange", "gray"))
abline(a = 15, b = 1, col = "orange")
abline(a = -15, b = 1, col = "orange")
```



```{r}
ggplot(data.frame(V, X), aes(x = V, y = X)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "white") +
  geom_abline(slope = 1, intercept = c(-15, 15), col = "orange", size = 2)
```






