# Homework 4 {.unnumbered}

{{< include _r_setup.qmd >}}



::: {#exm-binomial-sum}
Xavier bets on red on each of 3 spins of a roulette wheel.
After the third spin, Xavier leaves and Zander arrives and bets on black on the next two spins.
Let $X$ be the number of bets that Xavier wins and let $Z$ be the number that Zander wins.

1.  Identify the distribution of $X$ and the distribution of $Z$.
\
\
\
\
\
1.  Are $X$ and $Z$ independent?
Explain.
\
\
\
\
\
1.  Compute $\text{P}(X = 2, Z = 1)$. 
\
\
\
\
\
1.  What does $X+Z$ represent in this context?
Identify the distribution of $X+Z$ without doing any calculations.
\
\
\
\
\
1.  Compute $\text{P}(X + Z = 3)$. 
\
\
\
\
\
1.  Suppose that Yolanda bets on black on all 5 spins of the wheel.
Does $X + Y$ have a Binomial distribution?
Explain. 
\
\
\
\
\
1.  Suppose instead that Zavier bet on the number 7 in his two bets.
Does $X + Z$ have a Binomial distribution in this case?
Explain. 
\
\
\
\
\
:::


## Problem 6



Suppose the number of earthquakes per hour, for a certain range of magnitudes in a certain region, follows a Poisson distribution with parameter 0.7.

a. Compute and interpret the probability that there is at least one earthquake of this size in the region in any given hour.
a. Compute and interpret the probability that there are exactly 3 earthquakes of this size in the region in any given hour.
a. Interpret the value 0.7 in context.
a. Construct a table, plot, and spinner corresponding to a Poisson(0.7) distribution.


## Problem 1

In a certain region, times (minutes) between occurrences of earthquakes (of any magnitude) have a distribution with percentiles displayed in the table below.

a.  Construct a spinner corresponding to this distribution.
b.  What percent of times are between 26.8 and 110.0 minutes?
c.  Let $F$ be the cdf. Evaluate and interpret $F(42.8)$.
d.  Let $Q$ be the quantile function. Evaluate and interpret $Q(0.9)$.
e.  Sketch (by hand) a histogram of this distribution.

```{r}
#| echo: false
exponential_spinner <- data.frame(p = 0.1 * (1:9)) %>%
  mutate(x = round(qexp(p, rate = 1 / 120), 1))

exponential_spinner %>%
  mutate(p = paste(100 * p, "th", sep = "")) %>%
  kbl(col.names = c("Percentile", "Value (minutes)")) %>%
  kable_styling()
```

## Problem 2

(Continued from previous HW.) In the Regina/Cady problem, let $W=|R-Y|$ be the amount of time the first person to arrive has to wait for the second person.
Recall that $W$ is a continuous random variable with pdf $$
f_W(w) = 2(1-w), \quad 0 < w < 1.
$$ a.
Let $F$ be the cdf.
Evaluate and interpret $F(0.25)$.
a.
Find an expression for the cdf of $W$.
Set up an integral, but sketch a picture and use geometry to compute.
b.
Find and interpret the 25th percentile of $W$.
(You can do the next part first if you want, but it might help to start with a specific number like in this part.) c.
Find the quantile function of $W$.
d.
Sketch a spinner corresponding to the distribution of $W$.
Label the 25th, 50th, and 75th percentiles.
e.
**Coding required.** Use your simulation results from before to approximate the 25th, 50th, and 75th percentiles.

## Problem 3

Solve Example 11.3 from the Normal distribution handout (the problem on daily high temperatures in SLO.)

## Problem 4

Solve Example 11.4 from the Normal distribution handout (the problem on "effect size".)

## Problem 5

The latest series of collectible Lego Minifigures contains 3 different Minifigure prizes (labeled 1, 2, 3).
Each package contains a single unknown prize.
Suppose we only buy 3 packages and we consider as our sample space outcome the results of just these 3 packages (prize in package 1, prize in package 2, prize in package 3).
For example, 323 (or (3, 2, 3)) represents prize 3 in the first package, prize 2 in the second package, prize 3 in the third package.
Let $X$ be the number of distinct prizes obtained in these 3 packages.
Let $Y$ be the number of these 3 packages that contain prize 1.
Suppose that each package is equally likely to contain any of the 3 prizes, regardless of the contents of other packages.
There are 27 possible, equally likely outcomes

```{r}
#| echo: false
#| message: false

k = 3

outcomes = expand.grid(1:k, 1:k, 1:k)


n_outcomes = nrow(outcomes)

X = rep("", n_outcomes)
Y = rep("", n_outcomes)

outcomes <- outcomes %>%
  bind_cols(X, Y)

names(outcomes) = c(paste("box", 1:k, sep = ""), "X", "Y")

outcomes %>%
  kbl() %>%
  kable_styling(fixed_thead = TRUE)

```

a.  Evaluate $X$ and $Y$ for each of the outcomes.
b.  Construct a two-way table representing the *joint distribution* of $X$ and $Y$.
c.  Sketch a plot representing the joint distribution of $X$ and $Y$.
d.  Construct a spinner corresponding to the joint distribution of $X$ and $Y$.
e.  Describe two ways to simulate an $(X, Y)$ pair.
f.  Identify the *marginal* distribution of $X$, and construct a corresponding spinner.
g.  Identify the *marginal* distribution of $Y$, and construct a corresponding spinner.
h.  Describe in detail in words how you could conduct a simulation and use the results to approximate. (This part is asking you to describe the process in words; not to write code.)
    i. $\text{P}(X = 3)$
    i. $\text{P}(X = 2, Y = 1)$
    i. $\text{E}(X)$
    i. $\text{E}(Y)$
    i. $\text{E}(XY)$
    i. $\text{SD}(X)$
    i. $\text{SD}(Y)$
    1. $\text{Cov}(X, Y)$
    i. $\text{Corr}(X, Y)$
i.  **Coding required.** Code and run the simulation from the previous part and use the simulation results to approximate each of
    i. $\text{P}(X = 3)$
    i. $\text{P}(X = 2, Y = 1)$
    i. $\text{E}(X)$
    i. $\text{E}(Y)$
    i. $\text{E}(XY)$
    i. $\text{SD}(X)$
    i. $\text{SD}(Y)$
    1. $\text{Cov}(X, Y)$
    i. $\text{Corr}(X, Y)$
j.  Compute
    i. $\text{P}(X = 3)$
    i. $\text{P}(X = 2, Y = 1)$
    i. $\text{E}(X)$ iii.$\text{E}(Y)$
    i. $\text{E}(XY)$
    i. $\text{SD}(X)$ iii.$\text{SD}(Y)$
    1. $\text{Cov}(X, Y)$
    i. $\text{Corr}(X, Y)$


## Problem 2

In a certain population, household income $X$ (\$ thousands) follows the pdf
$$
f_X(x) = c x^{-2.5}, \quad x \ge 30
$$
for an appropriate constant $c$.
Note that the measurement units are \$ thousands, so 1 represents 1 thousand dollars.

a.  What are the possible values of $X$? Sketch the pdf of $X$.
b.  Find the value of $c$.
c.  Find the probability that a household has an income over \$200K.
d.  Find the probability that a household has an income exactly equal to \$200K.
e.  Without integrating, find the probability that a household has an income, rounded to the nearest thousand dollars, of \$200K.
f.  Without integrating, determine how many times more likely it is for a household to have an income, rounded to the nearest thousand dollars, of \$100K than \$200K.

## Problem 3

In the meeting time problem, assume that Regina's $R$ and Cady's $Y$ arrival times (measured in fractions of the hour after noon, e.g. 0.25 represents 12:15), each follow a Uniform(0, 1) distribution, independently of each other.
Let $W = |R - Y|$ be the amount of time (hours) the first person to arrive waits for the second person to arrive.
It can be shown that $W$ has pdf
$$
f_W(w) = 2(1-w), \qquad 0<w<1.
$$

a.  Sketch a plot of the pdf. Describe roughly what this means in context.
b.  **Coding required.** Code and run a simulation to approximate the distribution of $W$. Does the histogram reasonably match up with the pdf? Then use the simulation results to approximate $\text{P}(W < 0.25)$.
c.  Use the pdf to compute $\text{P}(W < 0.25)$ and interpret the value. Set up an integral, but sketch a picture and use geometry to compute.

## Problem 4

(Continued.) In the Regina/Cady problem, let $W=|R-Y|$ be the amount of time the first person to arrive has to wait for the second person.
Recall that $W$ is a continuous random variable with pdf
$$
f_W(w) = 2(1-w), \quad 0 < w < 1.
$$

a.  **Coding required.** Use your simulation results from before to approximate $\text{E}(W)$, $\text{E}(W^2)$, $\text{Var}(W)$, and $\text{SD}(W)$.
b.  Compute and interpret $\text{E}(W)$.
c.  Compute $\text{E}(W^2)$.
d.  Compute $\text{Var}(W)$
e.  Compute $\text{SD}(W)$.
f.  Explain in words in detail how you could use the Uniform(0, 1) spinner and simulation to approximate $\text{Var}(W)$.
g.  Compute the probability that $W$ is more than 1 SD away from its mean.


## Problem 5

Katniss throws a dart at a circular dartboard with radius 1 foot.
Suppose that Katniss's dart lands at a uniformly random location on the dartboard (and she never misses the dartboard).

a.  Compute the probability that Katniss's dart lands within 1 inch of the center of the dartboard.
b.  Compute the probability that Katniss's dart lands more than 1 inch but less than 2 inches away from the center of the dartboard.

