# Homework 6 solutions {.unnumbered}



{{< include _r_setup.qmd >}}

## Problem 1

For many animal species, the lengths of particular bones follow a Normal distribution.
For adult male humans, the length of the femur (thigh bone, the longest bone in humans) has an approximate Normal distribution with mean 48.2 cm and standard deviation of 4.5 cm.  
Complete each of the following parts by standardizing, drawing appropriate plots, and then providing an educated ballpark estimate based on the empirical rule. 
Then you can use software to get more precise values. 
But software should be your last step, not the first.

1.  What percent of adult male humans have a femur length above 50cm?
1.  What percent of adult male humans have a femur length between 40cm and 50cm?
1.  What is the 99th percentile of femur lengths for adult male humans?


### Solutions





1.  50 cm is (50-48.2)/4.5 = 0.4 SDs above the mean.  0.4 is close to 0.5; 38% of values are within 0.5 SDs of the mean, so 0.5 SDs above the mean is the 69th percentile.  So 31% of values are 0.5 or more SDs above the mean. Since 0.4 is less than 0.5 SDs above the mean, more than 31% of femur lengths will be greater than 50cm.  From software, we find 34.5%.
The z-score for 50 cm is (50-48.2)/4.5 = 0.4 SDs above the mean.  0.4 is close to 0.5; 38% of values are within 0.5 SDs of the mean, so 0.5 SDs above the mean is the 69th percentile.  Since 0.4 is less than 0.5 SDs above the mean, 

1.  From the previous part, less than 69% of femur lengths will be less than 50cm. 
From software, we find 65.5%. 
40cm is (40-48.2)/4.5 = -1.82, i.e. 1.82 SDs below the mean.  So a little more than more than 2.5% (from 2 SDs below the mean) of femur lengths will be less than 40cm.  From software we find 3.4%.  Subtracting shows 62.1% of femur lengths are between 40cm and 50cm.


1. From the empirical rule, we know that 97.5% of values are no more than 2 SDs above the mean, so we know the 99th percentile will be more than 2 SDs above the mean. Table or software shows that for a Normal distribution, the 99th percentile is 2.33 SDs above the mean.  So the 99th percentile for femur lengths is 48.2 + 2.33 × 4.5 = 58.7 cm.

 


## Problem 2

Daily high temperatures (degrees Fahrenheit) in San Luis Obispo in August follow (approximately) a Normal distribution with a mean of 76.9 degrees F.
The temperature exceeds 100 degrees Fahrenheit on about 1.5% of August days.

1.  Compute the standard deviation.
1.  Suppose the mean increases by 2 degrees Fahrenheit. 
On what percentage of August days will the daily high temperature exceed 100 degrees Fahrenheit? (Assume the standard deviation does not change.)
1.  A mean of 78.9 is 1.02 times greater than a mean of 76.9. 
By what (multiplicative) factor has the percentage of 100-degree days increased? What do you notice?


### Solution

1.  We are given that 100 is the 98.5th percentile.
Software shows that for a Normal distribution the 98.5th percentile is 2.17 standard deviations above the mean.
Set $100 = 76.9 + 2.17\sigma$ and solve for the standard deviation $\sigma = 10.65$ degrees F.

1.  The mean is now 78.9. Now the value 100 is $(100-78.9)/10.65 = 1.98$ standard deviations above the mean, around 2 so the probability we want is around 2.5%.
Software shows that for a Normal distribution the probability of a value more than 1.98 standard deviations above the mean is 0.023.
On 2.3% of August days the daily high temperature will exceed 100 degrees F.

1.  $0.023/0.015 = 1.53$. A 2 percent increase in the mean results in a 53 percent increase in the probability that a day has a daily high temperature above 100 degrees F.
Relatively small changes in the mean can have relatively large impacts on tail probabilities.



## Problem 3

The latest series of collectible Lego Minifigures contains 3 different Minifigure prizes (labeled 1, 2, 3).
Each package contains a single unknown prize.
Suppose we only buy 3 packages and we consider as our sample space outcome the results of just these 3 packages (prize in package 1, prize in package 2, prize in package 3).
For example, 323 (or (3, 2, 3)) represents prize 3 in the first package, prize 2 in the second package, prize 3 in the third package.
Let $X$ be the number of distinct prizes obtained in these 3 packages.
Let $Y$ be the number of these 3 packages that contain prize 1.
Suppose that each package is equally likely to contain any of the 3 prizes, regardless of the contents of other packages; let $\text{P}$ denote the corresponding probability measure.
There are 27 possible, equally likely outcomes

```{r}
#| echo: false
#| message: false

k <- 3

outcomes <- expand.grid(1:k, 1:k, 1:k)


n_outcomes <- nrow(outcomes)

X <- rep("", n_outcomes)
Y <- rep("", n_outcomes)

outcomes <- outcomes |>
  bind_cols(X, Y)

names(outcomes) <- c(paste("box", 1:k, sep = ""), "X", "Y")

outcomes |> kable()
```

1.  Evaluate $X$ and $Y$ for each of the outcomes.
1.  Construct a two-way table representing the joint distribution of $X$ and $Y$.
1.  Specify the marginal distribution of $X$ and compute $\text{E}(X)$.
1.  Specify the marginal distribution of $Y$ and compute $\text{E}(Y)$.
1.  Compute $\text{Cov}(X, Y)$
1.  Find the conditional distribution of $Y$ given $X=x$ for each possible value of $x$ of $X$.
1.  Compute and interpret $\text{E}(Y|X=x)$ for each possible value of $x$ of $X$.
1.  Find the conditional distribution of $X$ given $Y=y$ for each possible value of $y$ of $Y$.
1.  Compute and interpret $\text{E}(X|Y=y)$ for each possible value of $y$ of $Y$.
1.  Are $X$ and $Y$ independent? Explain; support your answer with appropriate numbers.



### Solution



```{r}
#| echo: false
#| warning: false
library(dplyr)
library(knitr)
library(kableExtra)

k <- 3

outcomes <- expand.grid(1:k, 1:k, 1:k)


n_outcomes <- nrow(outcomes)

X <- rep(NA, n_outcomes)
Y <- rep(NA, n_outcomes)

for (i in 1:n_outcomes) {
  X[i] <- outcomes %>% slice(i) %>% unlist() %>% n_distinct()
  Y[i] <- sum(outcomes[i, ] == 1)
}

outcomes <- outcomes %>%
  bind_cols(X, Y)

names(outcomes) <- c(paste("box", 1:k, sep = ""), "X", "Y")

outcomes %>%
  kbl() %>%
  kable_styling(fixed_thead = TRUE)
```


Two-way table: One dimension will have possible values of $x$, the other possible values of $y$.
The interior cells should contain the probability of each $(x, y)$ pair.

|       |      |       |      |      |       |
|:------|-----:|------:|-----:|-----:|------:|
|       |  $y$ |       |      |      |       |
| $x$   |    0 |     1 |    2 |    3 | Total |
| 1     | 2/27 |     0 |    0 | 1/27 |  3/27 |
| 2     | 6/27 |  6/27 | 6/27 |    0 | 18/27 |
| 3     |    0 |  6/27 |    0 |    0 |  6/27 |
| Total | 8/27 | 12/27 | 6/27 | 1/27 |     1 |


The possible values of $X$ are in the leftmost column and the marginal probabilities are in the Total column.
$E(X) = (1)(3/27)+(2)(18/27)+(3)(6/27) = 57/27 = 2.11$

The possible values of $Y$ are in the top row (0, 1, 2, 3) and the marginal probabilities are in the Total row.
$E(Y) = (0)(8/27)+(1)(12/27)+(2)(6/27)+(3)(1/27) = 1$


Covariance is 0. $E(XY) = (1)(0)(2/27) + (1)(3)(2/27)+\cdots+(3)(1)(6/27) = 57/27$. $E(XY)-E(X)E(Y)= (57/27)-(57/27)(1) = 0.$

Each row of the table below represents a different conditional distribution of $Y$ given $X=x$. For example, the row for $x=1$ represents the conditional distribution of $Y$ given $X=1$: Given $X=1$, $Y$ is 0 with probability 2/3 and 3 with probability 1/3.

|     |     |     |     |     |       |
|-----|----:|----:|----:|----:|------:|
|     | $y$ |     |     |     |       |
|     |   0 |   1 |   2 |   3 | Total |
| $x$ |     |     |     |     |       |
| 1   | 2/3 |   0 |   0 | 1/3 |     1 |
| 2   | 1/3 | 1/3 | 1/3 |   0 |     1 |
| 3   |   0 |   1 |   0 |   0 |     1 |


Take expected values according to each conditional distribution.  In general, $\text{E}(Y|X=x)$ depends on $x$, but in this case $\text{E}(Y|X=x) = 1$ for all values of $x$.; regardless of how many distinct prizez people obtain in their, the average number of prize 1s obtains is 1.

\begin{align*}
x & \quad \text{E}(Y|X=x)\\
1 & \quad 0(2/3) + 3(1/3) = 1\\
2 & \quad 0(1/3) + 1(1/3) + 2(1/3) = 1\\
3 & \quad 1(1) = 1
\end{align*}

Each column of the table below represents a different conditional distribution of $X$ given $Y=y$. For example, the column for $y=1$ represents the conditional distribution of $X$ given $Y=1$: Given $Y=1$, $X$ is 1 with probability 1/4 and 2 with probability 3/4.

|       |     |     |     |     |
|-------|----:|----:|----:|----:|
|       | $y$ |     |     |     |
|       |   0 |   1 |   2 |   3 |
| $x$   |     |     |     |     |
| 1     | 1/4 |   0 |   0 |   1 |
| 2     | 3/4 | 1/2 |   1 |   0 |
| 3     |   0 | 1/2 |   0 |   0 |
| Total |   1 |   1 |   1 |   1 |


Take expected values according to each conditional distribution. For example, $\text{E}(X|Y = 0) = 1.75$; among the people who never get prize 1 in their 3 boxes, the average number of distinct prizes they obtain is 1.75.

\begin{align*}
y & \quad \text{E}(X|Y=y)\\
0 & \quad 1(1/4) + 2(3/4) = 1.75\\
1 & \quad 2(1/2) + 3(1/2) = 2.5\\
2 & \quad 2(1) = 2\\
3 & \quad 1(1) = 1
\end{align*}

$X$ and $Y$ are NOT independent, despite the fact that their covariance is 0. The conditional distribution of $X$ changes based on the value of $Y$.

## Problem 4

Xavier and Yolanda are playing roulette.
They both place bets on red on the same 3 spins of the roulette wheel before Xavier has to leave.
(Remember, the probability that any bet on red on a single spin wins is 18/38.) 
After Xavier leaves, Yolanda places bets on red on 2 more spins of the wheel.
Let $X$ be the number of bets that Xavier wins (out of 3) and let $Y$ be the number that Yolanda wins (out of 5).

1.  Identify by name the marginal distribution of $X$. Be sure to specify the values of any relevant parameters. Compute $\text{E}(X)$.
1.  Identify by name the marginal distribution of $Y$. Be sure to specify the values of any relevant parameters. Compute $\text{E}(Y)$.
1.  The joint distribution of $X$ and $Y$ is represented in the table below. Explain why $p_{X, Y}(1, 4) = 0$ and $p_{X, Y}(2, 1) = 0$.
1.  Compute $p_{X, Y}(2, 3)$. (Yes, the table tells you it's 0.1766, but you have to show how this number can be computed based on the assumptions of the problem.)
1.  Compute $\text{Cov}(X, Y)$.
1.  Are $X$ and $Y$ independent? Explain; support your answer with appropriate numbers.


| $x$, $y$ |      0 |      1 |      2 |      3 |      4 |      5 |
|----------|-------:|-------:|-------:|-------:|-------:|-------:|
| 0        | 0.0404 | 0.0727 | 0.0327 |      0 |      0 |      0 |
| 1        |      0 | 0.1090 | 0.1963 | 0.0883 |      0 |      0 |
| 2        |      0 |      0 | 0.0981 | 0.1766 | 0.0795 |      0 |
| 3        |      0 |      0 |      0 | 0.0294 | 0.0530 | 0.0238 |


### Solution

1.  $X$ has a Binomial(3, 18/38) distribution so $\text{E}(X) = (3)(18/38)$.
1.  $Y$ has a Binomial(5, 18/38) distribution so $\text{E}(Y) = (5)(18/38)$.
1.  Since $X$ is the number of wins in the first 3 bets and $Y$ is the number of wins in the first 5 bets, $Y$ needs to be at least as big as $X$, and $Y$ cannot be bigger than $X+2$.
1.  $p_{X, Y}(2, 3)=P(X=2, Y=3)$. $X=2$ and $Y=3$ can only happen if there are 2 wins in the first 3 bets and 1 win in the last 2 bets. Since the bets are independent, the probability is $\left(\binom{3}{2}(18/38)^2(18/38)^1\right)\left(\binom{2}{1}(18/38)^1(18/38)^1\right)$
1.  $\text{Cov}(X, Y)=E(XY)-E(X)E(Y)=E(XY)-(3(18/38))(5(18/38)) = 0.75$. To find $E(XY)$ for each $(x, y)$ pair, find the product $xy$ then multiply by its probability and sum: $(0)(0)(0.0404) + (0)(1)(0.0727) + \cdots + (3)(5)(0.0238)$.
1.  The covariance is not 0, so $X$ and $Y$ are NOT independent.