# Homework 4 {.unnumbered}

{{< include _r_setup.qmd >}}


## Problem 1

In a certain region, times (minutes) between occurrences of earthquakes (of any magnitude) have a distribution with percentiles displayed in the table below.

a.  Construct a spinner corresponding to this distribution.
b.  What percent of times are between 26.8 and 110.0 minutes?
c.  Let $F$ be the cdf. Evaluate and interpret $F(42.8)$.
d.  Let $Q$ be the quantile function. Evaluate and interpret $Q(0.9)$.
e.  Sketch (by hand) a histogram of this distribution.

```{r}
#| echo: false
exponential_spinner = data.frame(p = 0.1 * (1:9)) %>%
  mutate(x = round(qexp(p, rate = 1 / 120), 1))

exponential_spinner %>%
  mutate(p = paste(100 * p, "th", sep = "")) %>%
  kbl(col.names = c("Percentile", "Value (minutes)")) %>%
  kable_styling()
  
  
  
```

## Problem 2

(Continued from previous HW.) In the Regina/Cady problem, let $W=|R-Y|$ be the amount of time the first person to arrive has to wait for the second person.
Recall that $W$ is a continuous random variable with pdf $$
f_W(w) = 2(1-w), \quad 0 < w < 1.
$$ a.
Let $F$ be the cdf.
Evaluate and interpret $F(0.25)$.
a.
Find an expression for the cdf of $W$.
Set up an integral, but sketch a picture and use geometry to compute.
b.
Find and interpret the 25th percentile of $W$.
(You can do the next part first if you want, but it might help to start with a specific number like in this part.) c.
Find the quantile function of $W$.
d.
Sketch a spinner corresponding to the distribution of $W$.
Label the 25th, 50th, and 75th percentiles.
e.
**Coding required.** Use your simulation results from before to approximate the 25th, 50th, and 75th percentiles.

## Problem 3

Solve Example 11.3 from the Normal distribution handout (the problem on daily high temperatures in SLO.)

## Problem 4

Solve Example 11.4 from the Normal distribution handout (the problem on "effect size".)

## Problem 5

The latest series of collectible Lego Minifigures contains 3 different Minifigure prizes (labeled 1, 2, 3).
Each package contains a single unknown prize.
Suppose we only buy 3 packages and we consider as our sample space outcome the results of just these 3 packages (prize in package 1, prize in package 2, prize in package 3).
For example, 323 (or (3, 2, 3)) represents prize 3 in the first package, prize 2 in the second package, prize 3 in the third package.
Let $X$ be the number of distinct prizes obtained in these 3 packages.
Let $Y$ be the number of these 3 packages that contain prize 1.
Suppose that each package is equally likely to contain any of the 3 prizes, regardless of the contents of other packages.
There are 27 possible, equally likely outcomes

```{r}
#| echo: false
#| message: false

k = 3

outcomes = expand.grid(1:k, 1:k, 1:k)


n_outcomes = nrow(outcomes)

X = rep("", n_outcomes)
Y = rep("", n_outcomes)

outcomes <- outcomes %>%
  bind_cols(X, Y)

names(outcomes) = c(paste("box", 1:k, sep = ""), "X", "Y")

outcomes %>%
  kbl() %>%
  kable_styling(fixed_thead = TRUE)

```

a.  Evaluate $X$ and $Y$ for each of the outcomes.
b.  Construct a two-way table representing the *joint distribution* of $X$ and $Y$.
c.  Sketch a plot representing the joint distribution of $X$ and $Y$.
d.  Construct a spinner corresponding to the joint distribution of $X$ and $Y$.
e.  Describe two ways to simulate an $(X, Y)$ pair.
f.  Identify the *marginal* distribution of $X$, and construct a corresponding spinner.
g.  Identify the *marginal* distribution of $Y$, and construct a corresponding spinner.
h.  Describe in detail in words how you could conduct a simulation and use the results to approximate. (This part is asking you to describe the process in words; not to write code.)
    i. $\text{P}(X = 3)$
    i. $\text{P}(X = 2, Y = 1)$
    i. $\text{E}(X)$
    i. $\text{E}(Y)$
    i. $\text{E}(XY)$
    i. $\text{SD}(X)$
    i. $\text{SD}(Y)$
    1. $\text{Cov}(X, Y)$
    i. $\text{Corr}(X, Y)$
i.  **Coding required.** Code and run the simulation from the previous part and use the simulation results to approximate each of
    i. $\text{P}(X = 3)$
    i. $\text{P}(X = 2, Y = 1)$
    i. $\text{E}(X)$
    i. $\text{E}(Y)$
    i. $\text{E}(XY)$
    i. $\text{SD}(X)$
    i. $\text{SD}(Y)$
    1. $\text{Cov}(X, Y)$
    i. $\text{Corr}(X, Y)$
j.  Compute
    i. $\text{P}(X = 3)$
    i. $\text{P}(X = 2, Y = 1)$
    i. $\text{E}(X)$ iii.$\text{E}(Y)$
    i. $\text{E}(XY)$
    i. $\text{SD}(X)$ iii.$\text{SD}(Y)$
    1. $\text{Cov}(X, Y)$
    i. $\text{Corr}(X, Y)$

