# Linear Combinations of Random Variables

{{< include _r_setup.qmd >}}

-   In general, the distribution of a sum (or linear combination) of random variables can be difficult to obtain and depends on the joint distribution of the random variables. 
-   But there are some special cases where the distribution of a sum (or linear combination) works out nicely. 
-   And there are shortcut formulas that allow us to find the expected value and variance of a sum (or linear combination) without first finding its distribution. 

```{r}
#| echo: false
#| warning: false
#| 
library(tidyverse)
library(knitr)
set.seed(21234)
```


::: {#exm-poisson-sum}
At a small hospital, let $X$ be the number of babies born in a 5 day work week and let $Y$ be the number of babies born on a weekend. 
Then $W = X + Y$ is the total number of babies born in a week. 
Assume that 

-   $X$ and $Y$ are independent
-   $X$ has a Poisson(2.1) distribution
-   $Y$ has a Poisson(0.4) distribution

@tbl-poisson-sum displays the marginal distributions of $X$ and $Y$. (The probabilities have been rounded a little.)
:::


|     x, y          |         0    |         1    |         2    |     P(X =   x)    |
|-------------------|-------------:|-------------:|-------------:|------------------:|
|     0             |              |              |              |          0.122    |
|     1             |              |              |              |          0.257    |
|     2             |              |              |              |          0.270    |
|     3             |              |              |              |          0.189    |
|     4             |              |              |              |          0.099    |
|     5             |              |              |              |          0.042    |
|     6             |              |              |              |          0.021    |
|     P(Y =   y)    |     0.670    |     0.268    |     0.062    |                   |

: Distributions in @exm-poisson-sum {#tbl-poisson-sum}

<!-- ::: {#tbl-poisson-sum layout-ncol=2}

|     x    |     P(X = x)    |
|----------|----------------:|
|     0    |        0.122    |
|     1    |        0.257    |
|     2    |        0.270    |
|     3    |        0.189    |
|     4    |        0.099    |
|     5    |        0.042    |
|     6    |        0.021    |

: Distribution of $X$ {#tbl-poisson-sum-x}

|     y    |     P(Y = y)    |
|----------|----------------:|
|     0    |        0.670    |
|     1    |        0.268    |
|     2    |        0.062    |

: Distribution of $Y$ {#tbl-poisson-sum-y}

Marginal distributions in @exm-poisson-sum.
::: -->


1.  Create a table for the joint distribution of $X$ and $Y$.
\
\
\
\
\
1.  Compute and interpret $\text{P}(W = 0)$.
\
\
\
\
\
1.  Compute and interpret $\text{P}(W = 1)$.
\
\
\
\
\
1.  Compute and interpret $\text{P}(W = 2)$.
\
\
\
\
\
1.  Find the distribution of $W$.
\
\
\
\
\
1.  Compute and interpret $\text{E}(W)$. How does it relate to $\text{E}(X)$ and $\text{E}(Y)$?
\
\
\
\
\
1.  Compute $\text{Var}(W)$. How does it relate to $\text{Var}(X)$ and $\text{Var}(Y)$?
\
\
\
\
\



-   **Poisson aggregation.** If $X$ and $Y$ are independent, $X$ has a Poisson($\mu_X$) distribution,  and $Y$ has a Poisson($\mu_Y$) distribution, then $X+Y$ has a Poisson($\mu_X+\mu_Y$) distribution.
-   If component counts are independent and each has a Poisson distribution, then the total count also has a Poisson distribution.



::: {#exm-sat-compare-ev}
We'll illustrate some ideas with a small sample of simulated values.
Say we have SAT Math scores ($X$) and Reading scores ($Y$) for 10 students. We're interested in the sum ($T = X + Y$) and difference (Math minus Reading, $D = X-Y$) of the scores.

The 10 $X$ values are the same in each scenario, and the 10 $Y$ values are the same in each scenario, but the $(X, Y)$ values are paired in different ways:

-   Scenario 1: correlation is 0.78
-   Scenario 2: correlation is about 0 (actually -0.04)
-   Scenario 3: correlation is -0.94

:::


| Student        |  $X$ | $Y$ |  $T$ | $D$ |
|----------------|-----:|----:|-----:|----:|
| 1              |  520 | 530 | 1050 | -10 |
| 2              |  540 | 470 | 1010 |  70 |
| 3              |  620 | 670 | 1290 | -50 |
| 4              |  620 | 630 | 1250 | -10 |
| 5              |  630 | 600 | 1230 |  30 |
| 6              |  670 | 610 | 1280 |  60 |
| 7              |  700 | 680 | 1380 |  20 |
| 8              |  700 | 580 | 1280 | 120 |
| 9              |  710 | 640 | 1350 |  70 |
| 10             |  760 | 720 | 1480 |  40 |
|                |      |     |      |     |
| Mean           |  647 | 613 | 1260 |  34 |
| SD             |   72 |  70 |  134 |  47 |
| Corr($X$, $Y$) | 0.78 |     |      |     |

: @exm-sat-compare-ev scenario 1, correlation is 0.78 {#tbl-sat-mean-table-positive-corr}


```{r}
#| label: fig-sat-mean-table-positive-corr
#| echo: false
#| fig-cap: "@exm-sat-compare-ev scenario 1, correlation is 0.78"
#| fig-subcap: 
#|   - "Scatterplot of $(X, Y)$"
#|   - "Rug plot of $X+Y$"
#|   - "Rug plot of $X-Y$"
#| layout-ncol: 3
x <- c(520, 540, 620, 620, 630, 670, 700, 700, 710, 760)
y <- c(530, 470, 670, 630, 600, 610, 680, 580, 640, 720)

# mean(x)
# mean(y)
# sx = sqrt(mean(x ^ 2) - mean(x) ^ 2); sx
# sy = sqrt(mean(y ^ 2) - mean(y) ^ 2); sy
# (mean(x * y) - mean(x) * mean(y)) / sx / sy
# sqrt(mean((x + y) ^ 2) - mean(x + y) ^ 2)
# sqrt(mean((x - y) ^ 2) - mean(x - y) ^ 2)

plot(x, y, xlim = c(500, 800), ylim = c(500, 800), xlab = "X", ylab = "Y")

ggplot(data.frame(x, y), aes(x = x + y)) +
  geom_rug(
    sides = "b",
    length = unit(14, "pt"),
    linewidth = 0.6,
    alpha = 1.0,
    color = "#1F78B4"
  ) +
  scale_x_continuous(
    limits = c(1000, 1500),
    breaks = seq(1000, 1500, 100),
    name = "X + Y"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(), # ← removes vertical grid lines
    axis.line.x = element_line(linewidth = 0.6, color = "black"),
    plot.margin = margin(10, 20, 20, 20)
  )

ggplot(data.frame(x, y), aes(x = x - y)) +
  geom_rug(
    sides = "b",
    length = unit(14, "pt"),
    linewidth = 0.6,
    alpha = 1.0,
    color = "#1F78B4"
  ) +
  scale_x_continuous(
    limits = c(-200, 300),
    breaks = seq(-200, 300, 100),
    name = "X - Y"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(), # ← removes vertical grid lines
    axis.line.x = element_line(linewidth = 0.6, color = "black"),
    plot.margin = margin(10, 20, 20, 20)
  )
```




| Student        |   $X$ | $Y$ |  $T$ |  $D$ |
|----------------|------:|----:|-----:|-----:|
| 1              |   520 | 680 | 1200 | -160 |
| 2              |   540 | 530 | 1070 |   10 |
| 3              |   620 | 630 | 1250 |  -10 |
| 4              |   620 | 600 | 1220 |   20 |
| 5              |   630 | 720 | 1350 |  -90 |
| 6              |   670 | 470 | 1140 |  200 |
| 7              |   700 | 670 | 1370 |   30 |
| 8              |   700 | 580 | 1280 |  120 |
| 9              |   710 | 640 | 1350 |   70 |
| 10             |   760 | 610 | 1370 |  150 |
|                |       |     |      |      |
| Mean           |   647 | 613 | 1260 |   34 |
| SD             |    72 |  70 |   98 |  103 |
| Corr($X$, $Y$) | -0.04 |     |      |      |


: @exm-sat-compare-ev scenario 2, correlation is about 0 {#tbl-sat-mean-table-zero-corr}


```{r}
#| label: fig-sat-mean-table-zero-corr
#| echo: false
#| fig-cap: "@exm-sat-compare-ev scenario 2, correlation is about 0"
#| fig-subcap: 
#|   - "Scatterplot of $(X, Y)$"
#|   - "Rug plot of $X+Y$"
#|   - "Rug plot of $X-Y$"
#| layout-ncol: 3
y <- c(680, 530, 630, 600, 720, 470, 670, 580, 640, 610)

# mean(x)
# mean(y)
# sx = sqrt(mean(x ^ 2) - mean(x) ^ 2); sx
# sy = sqrt(mean(y ^ 2) - mean(y) ^ 2); sy
# (mean(x * y) - mean(x) * mean(y)) / sx / sy
# sqrt(mean((x + y) ^ 2) - mean(x + y) ^ 2)
# sqrt(mean((x - y) ^ 2) - mean(x - y) ^ 2)

plot(x, y, xlim = c(500, 800), ylim = c(500, 800), xlab = "X", ylab = "Y")

ggplot(data.frame(x, y), aes(x = x + y)) +
  geom_rug(
    sides = "b",
    length = unit(14, "pt"),
    linewidth = 0.6,
    alpha = 1.0,
    color = "#1F78B4"
  ) +
  scale_x_continuous(
    limits = c(1000, 1500),
    breaks = seq(1000, 1500, 100),
    name = "X + Y"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(), # ← removes vertical grid lines
    axis.line.x = element_line(linewidth = 0.6, color = "black"),
    plot.margin = margin(10, 20, 20, 20)
  )

ggplot(data.frame(x, y), aes(x = x - y)) +
  geom_rug(
    sides = "b",
    length = unit(14, "pt"),
    linewidth = 0.6,
    alpha = 1.0,
    color = "#1F78B4"
  ) +
  scale_x_continuous(
    limits = c(-200, 300),
    breaks = seq(-200, 300, 100),
    name = "X - Y"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(), # ← removes vertical grid lines
    axis.line.x = element_line(linewidth = 0.6, color = "black"),
    plot.margin = margin(10, 20, 20, 20)
  )
```



| Student        |   $X$ | $Y$ |  $T$ |  $D$ |
|----------------|------:|----:|-----:|-----:|
| 1              |   520 | 720 | 1240 | -200 |
| 2              |   540 | 680 | 1220 | -140 |
| 3              |   620 | 670 | 1290 |  -50 |
| 4              |   620 | 640 | 1260 |  -20 |
| 5              |   630 | 630 | 1260 |    0 |
| 6              |   670 | 610 | 1280 |   60 |
| 7              |   700 | 600 | 1300 |  100 |
| 8              |   700 | 580 | 1280 |  120 |
| 9              |   710 | 530 | 1240 |  180 |
| 10             |   760 | 470 | 1230 |  290 |
|                |       |     |      |      |
| Mean           |   647 | 613 | 1260 |   34 |
| SD             |    72 |  70 |   26 |  140 |
| Corr($X$, $Y$) | -0.94 |     |      |      |

: @exm-sat-compare-ev scenario 3, correlation is -0.94 {#tbl-sat-mean-table-negative-corr}


```{r}
#| label: fig-sat-mean-table-negative-corr
#| echo: false
#| fig-cap: "@exm-sat-compare-ev scenario 3, correlation is -0.94"
#| fig-subcap: 
#|   - "Scatterplot of $(X, Y)$"
#|   - "Rug plot of $X+Y$"
#|   - "Rug plot of $X-Y$"
#| layout-ncol: 3
y <- c(720, 680, 670, 640, 630, 610, 600, 580, 530, 470)

# mean(x)
# mean(y)
# sx = sqrt(mean(x ^ 2) - mean(x) ^ 2); sx
# sy = sqrt(mean(y ^ 2) - mean(y) ^ 2); sy
# (mean(x * y) - mean(x) * mean(y)) / sx / sy
# sqrt(mean((x + y) ^ 2) - mean(x + y) ^ 2)
# sqrt(mean((x - y) ^ 2) - mean(x - y) ^ 2)

plot(x, y, xlim = c(500, 800), ylim = c(500, 800), xlab = "X", ylab = "Y")

ggplot(data.frame(x, y), aes(x = x + y)) +
  geom_rug(
    sides = "b",
    length = unit(14, "pt"),
    linewidth = 0.6,
    alpha = 1.0,
    color = "#1F78B4"
  ) +
  scale_x_continuous(
    limits = c(1000, 1500),
    breaks = seq(1000, 1500, 100),
    name = "X + Y"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(), # ← removes vertical grid lines
    axis.line.x = element_line(linewidth = 0.6, color = "black"),
    plot.margin = margin(10, 20, 20, 20)
  )

ggplot(data.frame(x, y), aes(x = x - y)) +
  geom_rug(
    sides = "b",
    length = unit(14, "pt"),
    linewidth = 0.6,
    alpha = 1.0,
    color = "#1F78B4"
  ) +
  scale_x_continuous(
    limits = c(-200, 300),
    breaks = seq(-200, 300, 100),
    name = "X - Y"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal(base_size = 14) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(), # ← removes vertical grid lines
    axis.line.x = element_line(linewidth = 0.6, color = "black"),
    plot.margin = margin(10, 20, 20, 20)
  )
```



In each scenario:

1.	Compute the mean of total scores.
\
\
\
\
\
2.	How does the mean of $X + Y$ relate to the means of $X$ and $Y$?
\
\
\
\
\
3.	Compute the mean of the difference in scores (Math – Reading)
\
\
\
\
\
4.	How does the mean of $X - Y$ relate to the means of $X$ and $Y$?
\
\
\
\
\



-   **Linearity of expected value.** For *any* two random variables $X$ and $Y$,
\begin{align*}
\text{E}(X + Y) & = \text{E}(X) + \text{E}(Y)
\end{align*}
-   That is, the expected value of the sum is the sum of expected values, regardless of how the random variables are related.
-   Therefore, you only need to know the marginal distributions of $X$ and $Y$ to find the expected value of their sum.  (But keep in mind that the *distribution* of $X+Y$ will depend on the joint distribution of $X$ and $Y$.)
-   Whether in the short run or the long run,
\begin{align*}
\text{Average of $X + Y$ } & = \text{Average of $X$} + \text{Average of $Y$}
\end{align*}
regardless of the joint distribution of $X$ and $Y$.
-   A **linear combination** of two random variables $X$ and $Y$ is of the form $aX + bY$ where $a$ and $b$ are non-random constants.  Combining properties of linear rescaling with linearity of expected value yields the expected value of a linear combination.
    $$
    \text{E}(aX + bY) = a\text{E}(X)+b\text{E}(Y)
    $$
-   For example ($a=1, b=-1$): $\text{E}(X-Y) = \text{E}(X) - \text{E}(Y)$.
-   Linearity of expected value extends naturally to more than two random variables.
    $$
    \text{E}(a_1X_1 + \cdots + a_nX_n) = a_1\text{E}(X_1) + \cdots + a_n\text{E}(X_n)
    $$

::: {#exm-sat-compare-var}
Continuing with the three scenarios in @exm-sat-compare-ev.

In each scenario:
:::

1.  Compute the variance of total scores.
\
\
\
\
\
1.  How does correlation affect $\text{Var}(X + Y)$? 
In which of the three scenarios is $\text{Var}(X + Y)$ the largest? 
The smallest?
Can you explain why? 
In which scenario is $\text{Var}(X + Y)$ roughly equal to the sum of $\text{Var}(X)$ and $\text{Var}(Y)$?
\
\
\
\
\
1.  Compute the variance of the difference in scores (Math - Reading)
\
\
\
\
\
1.  How does correlation affect $\text{Var}(X - Y)$? 
In which of the three scenarios is $\text{Var}(X - Y)$ the largest? 
The smallest?
Can you explain why? 
In which scenario is $\text{Var}(X - Y)$ roughly equal to the *sum* of $\text{Var}(X)$ and $\text{Var}(Y)$?
\
\
\
\
\

-   Variance of sums and differences of random variables
\begin{align*}
\text{Var}(X + Y) & = \text{Var}(X) + \text{Var}(Y) + 2\text{Cov}(X, Y)\\
\text{Var}(X - Y) & = \text{Var}(X) + \text{Var}(Y) - 2\text{Cov}(X, Y)
\end{align*}
-   If $X$ and $Y$ are uncorrelated:
\begin{align*}
\text{Var}(X+Y)  & = \text{Var}(X) + \text{Var}(Y)\qquad \text{if $X, Y$ are uncorrelated}\\
\text{Var}(X-Y)  & = \text{Var}(X) + \text{Var}(Y)\qquad \text{if $X, Y$ are uncorrelated}
\end{align*}
-   Variance of linear combinations: If $a, b, c$ are non-random constants and $X$ and $Y$ are random variables then
$$
\text{Var}(aX + bY + c) = a^2\text{Var}(X) + b^2\text{Var}(Y) + 2ab\text{Cov}(X, Y)
$$


::: {#exm-salty-beer}
Let $X$ be the sales of beer and $Y$ the sales of salty snacks in a week at a certain supermarket, measured in thousands of dollars.  
Suppose

-   $\text{E}(X) = 22$, $\text{SD}(X) = 6$
-   $\text{E}(Y) = 14$, $\text{SD}(Y) = 4$
-   $\text{Corr}(X, Y) = 0.55$
:::

1.  Compute and intepret $\text{E}(X+Y)$.
\
\
\
\
\
1.  Compute and intepret $\text{SD}(X+Y)$.
\
\
\
\
\
 

::: {#exm-wine-beer}
Let $X$ be the sales of beer and $Y$ the sales of wine in a week at a certain supermarket, measured in thousands of dollars.  
Suppose

-   $\text{E}(X) = 22$, $\text{SD}(X) = 6$
-   $\text{E}(Y) = 14$, $\text{SD}(Y) = 4$
-   $\text{Corr}(X, Y) = -0.55$
:::

1.  Compute and intepret $\text{E}(X+Y)$.
\
\
\
\
\
1.  Compute and intepret $\text{SD}(X+Y)$.
\
\
\
\
\




## Exercises

::: {#exr-var-two-times}
Consider a random variable $X$ with $\text{Var}(X)=1$. 
What is $\text{Var}(2X)$?

-   Walt says: $\text{SD}(2X) = 2\text{SD}(X)$ so $\text{Var}(2X) = 2^2\text{Var}(X) = 4(1) = 4$.
-   Jesse says: Variance of a sum is a sum of variances, so $\text{Var}(2X) = \text{Var}(X+X)$ which is equal to $\text{Var}(X)+\text{Var}(X) = 1+1=2$.

Who is correct? 
Why is the other wrong?
\
\
\
\
\
:::


::: {#exr-portfolio}
Percent returns for assets $X$, $Y$, and $Z$ follow a joint distribution with

-   Mean 10 and standard deviation 15 for asset $X$,
-   Mean 5 and standard deviation 3 for asset $Y$,
-   Correlation of −0.6 between asset $X$ and asset $Y$
-   Asset Z yields a constant return of 1 percent.

An investment portfolio has 60% of its funds in asset $X$, 30% in asset $Y$, and 10% in asset $Z$.

:::

1.  Let $R$ be the portfolio return. Express $R$ in terms of $X, Y, Z$.
\
\
\
\
\
1.  Compute $\text{E}(R)$.
\
\
\
\
\
1.  Compute $\text{Cov}(X, Y)$.
\
\
\
\
\
1.  Compute $\text{Cov}(X, Z)$.
\
\
\
\
\
1.  Compute $\text{SD}(R)$.