# Discrete Random Variables

{{< include _r_setup.qmd >}}

-   Roughly, a *random variable* assigns a number measuring some quantity of interest to each outcome of a random phenomenon.
-   Mathematically, a **random variable (RV)** $X$ is a *function* that takes an outcome in the sample space as input and returns a real number as output
-   The random variable itself is typically denoted with a capital letter ($X$); possible values of that random variable are denoted with lower case letters ($x$).
    -   Think of the capital letter $X$ as a label standing in for a formula like "the number of heads in 4 flips of a coin" and
    -   $x$ as a dummy variable standing in for a particular value like 3.
-   **Discrete random variables** take at most countably many possible values (e.g., $0, 1, 2, \ldots$). They are often counting variables (e.g., the number of Heads in 10 coin flips).
-   **Continuous random variables** can take any real value in some interval (e.g., $[0, 1]$, $[0,\infty)$, $(-\infty, \infty)$.). That is, continuous random variables can take uncountably many different values. Continuous random variables are often measurement variables (e.g., height, weight, income).
-   A function of a random variable is also a random variable: if $X$ is a RV then so is $g(X)$
-   Sums and products, etc., of random variables *defined on the same sample space* are random variables. If$X$ and $Y$ are RVs defined on the same sample space then so are $X+Y$, $X-Y$, $XY$
-   If the sample space outcomes are represented by rows in a spreadsheet, then random variables correspond to columns.
-   Expressions like $X=x$ or $\{X=x\}$ represent *events*: for which outcomes is the value of the random variable $X$ equal to the value $x$. (Remember, if the sample space outcomes are represented by rows in a spreadsheet, then an event corresponds to a subset of rows (outcomes) that satisfies some criteria.)
-   The **(probability) distribution** of a collection of random variables identifies the possible values that the random variables can take and their relative likelihoods.
-   We will see many ways of describing a distribution, depending on how many random variables are involved and their types (discrete or continuous).

```{r}
#| label: tbl-dice-rv
#| echo: false
#| warning: false
#| tbl-cap: "Table representing the sum ($X$) and larger ($Y$) of two rolls of a four-sided die"

u1 <- sort(rep(1:4, 4))
u2 <- rep(1:4, 4)
x <- u1 + u2
y <- pmax(u1, u2)

kbl(
  data.frame(paste("(", u1, ", ", u2, ")", sep = ""), x, y),
  col.names = c("Outcome (First roll, second roll)", "X (sum)", "Y (max)")
) |>
  kable_styling(fixed_thead = TRUE)
```

::: {#exm-dice-rv}
Roll a four-sided die twice, and record the result of each roll in sequence.
Let $X$ be the sum of the two dice, and let $Y$ be the larger of the two rolls (or the common value if both rolls are the same).
@tbl-dice-rv represents the possible outcomes and the values of the random variables $X$ and $Y$.
Assume the die is fair and the rolls are independent.
:::

1.  Identify the event $\{X = 4\}$ and compute its probability. Then interpret the probability both as a long relative frequency and as a relative likelihood.\
    \
    \
    \
    \
2.  Construct a table and plot of $\text{P}(X = x)$ for each possible value $x$ of $X$.\
    \
    \
    \
    \
3.  Construct a table and plot of $\text{P}(Y = y)$ for each possible value $y$ of $Y$.\
    \
    \
    \
    \

```{r}
#| label: tbl-dice-marginal
#| echo: false
#| tbl-cap: 
#|   - "The marginal distribution of $X$, the sum of two rolls of a fair four-sided die."
#|   - "The marginal distribution of $Y$, the larger of two rolls of a fair four-sided die."
#| layout-ncol: 2

x <- 2:8
px <- c(1, 2, 3, 4, 3, 2, 1) / 16

kbl(
  data.frame(x, px),
  col.names = c("x", "P(X=x)"),
  digits = 4
) |>
  kable_styling(fixed_thead = TRUE)

y <- 1:4
py <- c(1, 3, 5, 7) / 16

knitr::kable(
  data.frame(y, py),
  col.names = c("y", "P(Y=y)"),
  digits = 4
) |>
  kable_styling(fixed_thead = TRUE)
```

```{r}
#| label: fig-dice-marginal
#| echo: false
#| fig-cap: 
#|   - "The marginal distribution of $X$, the sum of two rolls of a fair four-sided die."
#|   - "The marginal distribution of $Y$, the larger of two rolls of a fair four-sided die."
#| layout-ncol: 2

ggplot(data.frame(x, px), aes(x = x, xend = x, y = 0, yend = px)) +
  geom_segment(linewidth = 1.2, col = "skyblue") +
  scale_x_continuous(breaks = 2:8) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.5)) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "x", y = "P(X = x)")

ggplot(data.frame(y, py), aes(x = y, xend = y, y = 0, yend = py)) +
  geom_segment(linewidth = 1.2, col = "orange") +
  scale_x_continuous(breaks = 2:8) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.5)) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "y", y = "P(Y = y)")
```

## Simulating from a Distribution

::: {#exm-dice-spinners}
Continuing @exm-dice-rv.
:::

1.  Describe how you could simulate a single value of $Y$.\
    \
    \
    \
    \
2.  Construct a spinner (like from a kids game) to represent the distribution of $Y$.\
    \
    \
    \
    \
3.  Describe another way to simulate a single value of $Y$.\
    \
    \
    \
    \
4.  Describe how you could use simulation to approximate the distribution of $Y$.\
    \
    \
    \
    \

```{r}
#| label: fig-dice-marginal-spinner
#| echo: false
#| fig-cap: 
#|   - "Spinner representing the marginal distribution of $X$, the sum of two rolls of a fair four-sided die."
#|   - "Spinner representing the marginal distribution of $Y$, the larger of two rolls of a fair four-sided die."
#| layout-ncol: 2

n <- 7

xp <- data.frame(
  x = 2:8,
  p = (4 - abs(2:8 - 5)) / 16
)

cdf <- c(0, cumsum(xp$p))

plotp <- (cdf[-1] + cdf[-length(cdf)]) / 2

ggplot(xp, aes(x = "", y = p, fill = x)) +
  geom_bar(width = 1, stat = "identity", color = "black", fill = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = plotp, labels = xp$x) +
  theme(axis.text.x = element_text(size = 14, face = "bold")) +
  # plot the probabilities as percents inside
  geom_text(aes(y = plotp, label = percent(p, accuracy = 0.01)), size = 3)


xp <- data.frame(
  x = 1:4,
  p = (2 * (1:4) - 1) / 16
)

cdf <- c(0, cumsum(xp$p))

plotp <- (cdf[-1] + cdf[-length(cdf)]) / 2

ggplot(xp, aes(x = "", y = p, fill = x)) +
  geom_bar(width = 1, stat = "identity", color = "black", fill = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = plotp, labels = xp$x) +
  theme(axis.text.x = element_text(size = 14, face = "bold")) +
  # plot the probabilities as percents inside
  geom_text(aes(y = plotp, label = percent(p, accuracy = 0.01)), size = 3)
```

-   Don't confuse a random variable with its distribution!
    -   A random variable measures a numerical quantity which depends on the outcome of a random phenomenon
    -   The distribution of a random variable specifies the long run pattern of variation of values of the random variable over many repetitions of the underlying random phenomenon.
-   Any marginal distribution can be represented by a single spinner (like from a kids game).
-   In principle, there are always two ways of simulating a value $x$ of a random variable $X$.
    1.  *Simulate from the probability space.* Simulate an outcome $\omega$ from the underlying probability space and set $x = X(\omega)$.
    2.  *Simulate from the distribution.* Construct a spinner corresponding to the distribution of $X$ and spin it once to generate $x$.
-   The second method requires that the distribution of $X$ is known. However, as we will see in many examples, *it is common to specify the distribution of a random variable directly without defining the underlying probability space*.

## Probability Mass Functions

-   The probability distribution of a single discrete random variable $X$ is often displayed in a table containing the probability of the event $\{X=x\}$ for each possible value $x$.
-   In some cases, a distribution has a "formulaic" shape. For a discrete random variable $X$, the **probability mass function (pmf)** $p_X$ expresses $\text{P}(X=x)$ as a function of $x$: $p_X(x) = \text{P}(X = x)$.
-   Be sure to specify the possible values of the random variable!
-   Certain common distributions have special names and properties.

::: {#exm-dice-pmf}
Continuing @exm-dice-rv.
:::

1.  Verify that the pmf of $X$ is $$
    p_X(x) = \frac{4 - |x - 5|}{16}, \qquad x = 2, 3, \ldots, 8
    $$\
    \
    \
    \
    \
2.  Verify that the pmf of $Y$ is $$
    p_Y(y) = \frac{2y - 1}{16}, \qquad y = 1, 2, 3, 4
    $$\
    \
    \
    \
    \

## Matching Problem

The "matching problem" involves $n$ distincts objects labeled $1, \ldots, n$ which are placed in $n$ distinct boxes labeled $1, \ldots, n$, with exactly one object placed in each box.
Suppose the objects are placed in the boxes uniformly at random, so that any possible arrangement is equally likely.
Let $X$ be the number of matches, that is, the number of objects for which their label matches the label of the box in which they are placed.

```{r}
#| label: tbl-matching-outcome
#| echo: false
#| tbl-cap: "Table representing the sample space in the matching problem with $n=4$."

n <- 4

permutations(n, n, 1:n, repeats.allowed = FALSE) |>
  as.data.frame() |>
  add_column(X = "") |>
  kbl(col.names = c(paste("Spot ", 1:n), "X (number of matches)")) |>
  kable_styling(fixed_thead = TRUE)
```

::: {#exm-matching-pmf}
Consider the matching problem with $n=4$.
There are $4!=24$ possible outcomes, displayed in @tbl-matching-outcome.
:::

1.  Complete @tbl-matching-outcome to identify the value of $X$ for each outcome.\
    \
    \
    \
    \
2.  Identify the event $\{X = 0\}$ and compute its probability. Then interpret the probability both as a long relative frequency and as a relative likelihood.\
    \
    \
    \
    \
3.  Construct a table, plot, and spinner representing the distribution of $X$.\
    \
    \
    \
    \

```{r}
#| label: tbl-matching-dist
#| echo: false
#| tbl-cap: "Theoretical distribution of $X$ (number of matches) in the matching problem with $n=4$."

kbl(
  data.frame(c(0, 1, 2, 4), round(c(9, 8, 6, 1) / 24, 4)),
  booktabs = TRUE,
  col.names = c("x", "P(X=x)")
) %>%
  kable_styling(fixed_thead = TRUE)
```

```{r}
#| label: fig-matching-dist-spinner
#| echo: false
#| fig-cap: "Spinner representing the theoretical distribution of $X$, the number of matches in the matching problem when $n=4$."

x <- c(0:2, 4)
p <- c(9, 8, 6, 1) / 24

xp <- data.frame(x, p)

cdf <- c(0, cumsum(xp$p))

plotp <- (cdf[-1] + cdf[-length(cdf)]) / 2

ggplot(xp, aes(x = "", y = p, fill = x)) +
  geom_bar(width = 1, stat = "identity", color = "black", fill = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = plotp, labels = xp$x) +
  theme(axis.text.x = element_text(size = 14, face = "bold")) +
  # plot the probabilities as percents inside
  geom_text(
    aes(y = plotp, label = percent(p)),
    size = 3,
    color = c(rep("black", 4))
  )
```



Randomly select a county in the U.S.
Let $X$ be the leading digit in the county's population.
[For example](https://en.wikipedia.org/wiki/List_of_counties_in_California), if the county's population is 10,040,000 (Los Angeles County) then $X=1$; if 3,170,000 (Orange County) then $X=3$; if 283,000 (SLO County) then $X=2$; if 30,600 (Lassen County) then $X=3$.
The possible values of $X$ are $1, 2, \ldots, 9$.
You might think that $X$ is equally likely to be any of its possible values.
However, a more appropriate model is to assume that $X$ has pmf
$$
p_X(x) = 
\begin{cases}
\log_{10}(1+\frac{1}{x}), & x = 1, 2, \ldots, 9,\\
0, & \text{otherwise}
\end{cases}
$$
This distribution is known as [Benford's law](https://en.wikipedia.org/wiki/Benford's_law).

a.  Construct a table specifying the distribution of $X$, and the corresponding spinner.\
a.  Find $\text{P}(X \ge 3)$\
a.  **Coding required.** Code and run a simulation and use the results to approximate the distribution of $X$. (In R, there are built in functions to simulate Benford's law, but I want you to use the `sample` function. In R, `log` is natural log; `log10` is base-10 log.)
   


::: {#exm-matching-poisson}
Now consider the matching problem with general $n$.
Finding the distribution of $X$ by enumerating the $n!$ possible outcomes as we did for $n=4$ is not feasible for general $n$.
But we can approximate the distribution of $X$ with simulation.
Simulation shows that the distribution of the number of matches $X$ is approximately the same for all $n$ (unless $n$ is very small, i.e., $n\le 5$).
In particular, for any $n$, the probability mass function of $X$ is approximately
$$
p_X(x) = \frac{e^{-1}}{x!}, \qquad x = 0, 1, 2, \ldots
$$
:::

1.  Use this pmf to approximate the probability of at least one match, and compare to the simulation results for general $n$.\
    \
    \
    \
    \
2.  Interpret the probability both as a long relative frequency and as a relative likelihood.\
    \
    \
    \
    \
3.  Construct a table, plot, and spinner corresponding to the above pmf. Compare to the simulation results for general $n$.\
    \
    \
    \
    \

```{r}
#| label: tbl-poisson-matching
#| echo: false
#| tbl-cap: "Approximate distribution of number of matches in the matching problem for general $n$"

x <- 0:7

p_x <- exp(-1) / factorial(x)

data.frame(x, p_x) |>
  kbl(digits = 6) |>
  kable_styling(fixed_thead = TRUE)
```

```{r}
#| label: fig-poisson-matching-plot
#| echo: false
#| fig-cap: "Approximate distribution of number of matches in the matching problem for general $n$"
 
plot(
  x,
  p_x,
  type = "h",
  xlab = "Number of matches (x)",
  ylab = "Approximate probability p(x)"
)
```

```{r}
#| label: fig-poisson-matching-spinner
#| echo: false
#| fig-cap: "Spinner representing the approximate distribution of number of matches in the matching problem for general $n$"

x <- c(0:3, "4+")
p <- c(dpois(0:3, 1), 1 - ppois(3, 1))

xp <- data.frame(x, p)

cdf <- c(0, cumsum(xp$p))

plotp <- (cdf[-1] + cdf[-length(cdf)]) / 2

ggplot(xp, aes(x = "", y = p, fill = x)) +
  geom_bar(width = 1, stat = "identity", color = "black", fill = "white") +
  coord_polar("y", start = 0) +
  theme_void() +
  # plot the possible values on the outside
  scale_y_continuous(breaks = plotp, labels = xp$x) +
  theme(axis.text.x = element_text(size = 14, face = "bold")) +
  # plot the probabilities as percents inside
  geom_text(
    aes(y = plotp, label = percent(p)),
    size = 3,
    color = c(rep("black", 4), rep("white", 1))
  )
```

